---
title: "GLMM"
format: html
cache: TRUE
warning: FALSE
---
## Overview
In this analysis we will be running gamma regressions on our avian data.
This analysis includes frameworks for analysis and testing on:

-  Base model with our data

-  Sample method with our data

-  Adjusted model with our data

Load Packages, functions, and data
```{r}
library(tidyverse) #2.0.0
library(sf) #1.0-21
library(here) #1.0.1
library(janitor) #2.2.1
library(terra) #1.8-42
library(glmmTMB) #1.1.14
library(broom.mixed) #0.2.9.6
library(gtsummary) #2.5.0
library(gtExtras) #0.6.2
library(lmtest) #0.9-40
library(spdep) #1.4-2
library(kableExtra) #1.4.0
library(sfarrow) #0.4.1
library(tigris) #2.2.1
library(webshot2) #0.1.2


source(here('R','fit_agg.R'))
source(here('R','fit_disagg.R'))
source(here('R','adj_fit_agg.R'))
source(here('R', 'adj_fit_disagg.R'))
source(here('R', 'autocorrelation_tests.R'))

source(here('R','point_blue_theme.R'))

set.seed(76)


birds_joined <- st_read_parquet(here('data', 'data_processed', 'birds_joined.parquet'))

habitats <- unique(birds_joined$habitat_type)

```


## Base Models
We first tested our base models and assessed for temporal and spatial autocorrelation. With more data the base models may show no temporal or spatial autocorrelation.
```{r}
#| warning: FALSE

# Create data frame to be used in model
biodiv_area_search_agg <- birds_joined %>% 
  # Filter for survey type to separate them
    filter(survey_type %in% "Area Search")  %>%
  # Arrange for durbin watson
    arrange(desc(observation_date))
  
# Run data through model
  area_agg <- fit_agg(biodiv_area_search_agg)
  
# Repeat for aggregated, disagregated
  biodiv_point_count_agg <- birds_joined %>% 
    filter(survey_type %in% "Point Count") %>% 
    arrange(desc(observation_date))
  
  point_agg <- fit_agg(biodiv_point_count_agg)
  
  biodiv_area_search_disag <- birds_joined %>% 
    mutate(gap_sts = as.numeric(gap_sts)) %>% 
    arrange(desc(observation_date))
  
  area_disag <- fit_disagg(biodiv_area_search_disag) 
  
  biodiv_point_count_disag <- birds_joined %>% 
    mutate(gap_sts = as.numeric(gap_sts)) %>% 
    filter(survey_type %in% "Point Count") %>% 
    arrange(desc(observation_date))
  
 point_disag <- fit_disagg(biodiv_point_count_disag)
  
  
 # This is for a general summary (not separating by survey type)
  birds_rich <- birds_joined %>%  
    arrange(desc(observation_date))
  
  bird_summary <- fit_agg(birds_rich)
  

# Aggregate models for the function
models <- list(
  point_disag  = point_disag,
  area_disag   = area_disag,
  point_agg    = point_agg,
  area_agg     = area_agg,
  bird_summary = bird_summary
)

spatial_temporal_tests <- list()

# Aggregate data for morans
model_data_map <- list(
  point_disag  = biodiv_point_count_disag,
  area_disag   = biodiv_area_search_disag,
  point_agg    = biodiv_point_count_agg,
  area_agg     = biodiv_area_search_agg,
  bird_summary = birds_rich
)

# Run function
spatial_temporal_tests <- run_spatial_temporal_tests(
  models         = models,
  model_data_map = model_data_map,
  habitats       = habitats
)
# Print to assess (can be overwhelming)
print(spatial_temporal_tests, n = 38)
```

## Base model kable tables 

These kable tables are used to assess the models and were primarily used in the technical memorandum. Currently they save to our `outputs/figures` but the last line can be deleted to display locally

```{r}
#| warning: FALSE


## Aggregated Durbin Watson 
spatial_temporal_tests %>%
  filter(model_name %in% c('point_agg', 'area_agg')) %>%
  select(model_name, habitat, dw) %>%
  mutate(habitat = str_to_title(habitat),
         dw = round(dw, 3)) %>%
  pivot_wider(names_from = model_name, values_from = dw) %>%
  rename('Type' = habitat, 
         'Point Count' = 'point_agg', 
         'Area Search' = 'area_agg') %>%
   kable() %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  add_header_above(c("Habitat" = 1, "Durbin-Watson" = 2), 
                   background = "#4A90C4", color = "white", bold = TRUE) %>%
  add_header_above(c('Aggregated Durbin-Watson Statistics by Survey Method' = 3),  background = "#4A90C4", color = "white", bold = TRUE) %>% 
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, 7, 2), background = "#8EBEE2") %>%
  row_spec(seq(2, 6, 2), background = "#D6E8F5") %>%
  footnote(general = "DW values near 2 indicate no temporal autocorrelation.") %>% 
  save_kable(here('Outputs', 'figures', "agg_dw.png"), zoom = 3)

## Disaggregated Durbin Datson 
spatial_temporal_tests %>%
  filter(model_name %in% c('point_disag', 'area_disag')) %>%
  select(model_name, habitat, dw) %>%
  mutate(habitat = str_to_title(habitat),
         dw = round(dw,3)) %>%
  pivot_wider(names_from = model_name, values_from = dw) %>%
  rename('Type' = habitat, 
         'Point Count' = point_disag, 
         'Area Search' = area_disag) %>%
   kable() %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  add_header_above(c("Habitat" = 1, "Durbin-Watson" = 2), 
                   background = "#4A90C4", color = "white", bold = TRUE) %>%
  add_header_above(c('Disaggregated Durbin-Watson Statistics by Survey Method' = 3),  background = "#4A90C4", color = "white", bold = TRUE) %>% 
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, 7, 2), background = "#8EBEE2") %>%
  row_spec(seq(2, 6, 2), background = "#D6E8F5") %>%
  footnote(general = "DW values near 2 indicate no autocorrelation.") %>% 
  save_kable(here('Outputs', 'figures',"disag_dw.png"), zoom = 3)

## Aggregated Moran's Index
spatial_temporal_tests %>%
  filter(model_name %in% c('point_disag', 'area_disag')) %>%
  select(model_name, habitat, moran_I) %>%
  mutate(habitat = str_to_title(habitat),
         moran_I = round(moran_I, 3)) %>%
  pivot_wider(names_from = model_name, values_from = moran_I) %>%
  rename('Type' = habitat, 
         'Point Count' = point_disag, 
         'Area Search' = area_disag) %>%
   kable() %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  add_header_above(c("Habitat" = 1, "Moran's Index" = 2), 
                   background = "#4A90C4", color = "white", bold = TRUE) %>%
  add_header_above(c("Disaggregated Moran's Index Statistics by Survey Method" = 3),  background = "#4A90C4", color = "white", bold = TRUE) %>% 
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, 7, 2), background = "#8EBEE2") %>%
  row_spec(seq(2, 6, 2), background = "#D6E8F5") %>%
  footnote(general = "Moran's Index values near 0 indicate no spatial autocorrelation.")%>% 
  save_kable(here('Outputs', 'figures',"disag_moran.png"), zoom = 3)

## Disaggregated Moran's Index
spatial_temporal_tests %>%
  filter(model_name %in% c('point_agg', 'area_agg')) %>%
  select(model_name, habitat, moran_I) %>%
  mutate(habitat = str_to_title(habitat),
         moran_I = round(moran_I, 3)) %>%
  pivot_wider(names_from = model_name, values_from = moran_I) %>%
  rename('Type' = habitat, 
         'Point Count' = 'point_agg', 
         'Area Search' = 'area_agg') %>%
   kable() %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  add_header_above(c("Habitat" = 1, "Moran's Index" = 2), 
                   background = "#4A90C4", color = "white", bold = TRUE) %>%
  add_header_above(c("Aggregated Moran's Index Statistics by Survey Method" = 3),  background = "#4A90C4", color = "white", bold = TRUE) %>% 
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, 7, 2), background = "#8EBEE2") %>%
  row_spec(seq(2, 6, 2), background = "#D6E8F5") %>%
  footnote(general = "Moran's Index values near 0 indicate no spatial autocorrelation.") %>% 
  save_kable(here('Outputs', 'figures',"agg_moran.png"), zoom = 3)
```

## Sample Models 

Next we assessed sample models to see if it improves spatial autocorrelation. We create a grid and sample one point from each grid cell.

```{r}
#| warning: FALSE

# Grab california boundary from tigris package
ca_boundary <- states( cb = TRUE, class = "sf") %>% 
  filter(STUSPS == 'CA') %>% 
  st_transform(crs = st_crs(birds_joined))

# Create grid
grid_sf <- ca_boundary %>% 
  st_make_grid(cellsize = units::set_units(20,km)) %>% 
  st_intersection(ca_boundary) %>% 
  # st_intersection returns geometry column, 
  # convert to sf so it could store cell_id attribute
  st_as_sf() %>% 
  st_set_geometry("geometry") %>% 
  # factor : for more convenient ggplot viz 
  mutate(cell_id = as.factor(row_number())) %>% 
  st_transform(crs = "WGS84")

# Make sampled data frame with one sample per cell and year collected
sample_biodiv_area_search_agg <- birds_joined %>% 
    filter(survey_type %in% "Area Search")  %>% 
    st_transform(crs = "WGS84") %>% 
    st_join(grid_sf, left = FALSE) %>% 
    slice_sample(n = 1, by = c(cell_id, year_collected))
  
# Run base model on sampled data
  sample_area_agg <- fit_agg(sample_biodiv_area_search_agg)
  
  sample_biodiv_point_count_agg <- birds_joined %>% 
    filter(survey_type %in% "Point Count") %>% 
    st_transform(crs = "WGS84") %>% 
    st_join(grid_sf, left = FALSE) %>% 
    slice_sample(n = 1, by = c(cell_id, year_collected))
  
  sample_point_agg <- fit_agg(sample_biodiv_point_count_agg)
  
  sample_biodiv_area_search_disag <- birds_joined %>% 
    mutate(gap_sts = as.factor(gap_sts)) %>% 
    filter(survey_type %in% "Area Search") %>%
    st_transform(crs = "WGS84") %>% 
    st_join(grid_sf, left = FALSE) %>% 
    slice_sample(n = 1, by = c(cell_id, year_collected))
  
  sample_area_disag <- fit_disagg(sample_biodiv_area_search_disag) 
  
  sample_biodiv_point_count_disag <- birds_joined %>% 
    mutate(gap_sts = as.factor(gap_sts)) %>% 
    filter(survey_type %in% "Point Count") %>% 
    st_transform(crs = "WGS84") %>% 
    st_join(grid_sf, left = FALSE) %>% 
    slice_sample(n = 1, by = c(cell_id, year_collected))
  
 sample_point_disag <- fit_disagg(sample_biodiv_point_count_disag)
  

models <- list(
  sample_point_disag  = sample_point_disag,
  sample_area_disag   = sample_area_disag,
  sample_point_agg    = sample_point_agg,
  sample_area_agg     = sample_area_agg
)

sample_spatial_temporal_tests <- list()

sample_model_data_map <-  list()

sample_model_data_map <- list(
  sample_point_disag  = sample_biodiv_point_count_disag,
  sample_area_disag   = sample_biodiv_area_search_disag,
  sample_point_agg    = sample_biodiv_point_count_agg,
  sample_area_agg     = sample_biodiv_area_search_agg
)


sample_spatial_temporal_tests <- run_spatial_temporal_tests(
  models         = models,
  model_data_map = sample_model_data_map,
  habitats       = habitats
)

sample_spatial_temporal_tests
```

When you examine this, it does not really improve spatial or temporal autocorrelation. However with more computing power you can adjust the grid size and potentially find different results. 

## Adjust models

Next, we adjusted the models. This ended up improving the autocorrelation.

```{r}
#| warning: FALSE
# Here we do not need to update the data

adj_area_agg <- adj_fit_agg(biodiv_area_search_agg)

adj_point_agg <- adj_fit_agg(biodiv_point_count_agg)

adj_area_disag <- adj_fit_disagg(biodiv_area_search_disag) 

adj_point_disag <- adj_fit_disagg(biodiv_point_count_disag)

adj_bird_summary <- adj_fit_agg(birds_rich)

models <- list(
  adj_point_disag  = adj_point_disag,
  adj_area_disag   = adj_area_disag,
  adj_point_agg    = adj_point_agg,
  adj_area_agg     = adj_area_agg
)

adj_spatial_temporal_tests <- list()

model_data_map <- list(
  adj_point_disag  = biodiv_point_count_disag,
  adj_area_disag   = biodiv_area_search_disag,
  adj_point_agg    = biodiv_point_count_agg,
  adj_area_agg     = biodiv_area_search_agg
)


adj_spatial_temporal_tests <- run_spatial_temporal_tests(
  models         = models,
  model_data_map = model_data_map,
  habitats       = habitats
)

print(adj_spatial_temporal_tests, n=38)
```

# Create model summary of aggregated biodiversity across point and area search

```{r}
#| warning: FALSE

birds_rich <- birds_joined %>% 
  # Arrange by observation date descending
  arrange(desc(observation_date))

habitats <- unique(birds_rich$habitat_type)
tmp <- vector(mode = 'list', length = 0)
bird_summary <- vector(mode = 'list', length = 0)

for (i in habitats){
  
  tmp[[i]] <- birds_rich %>% 
    st_drop_geometry() %>% 
    filter(habitat_type == i)
  
  
  bird_summary[[i]] <- glmmTMB(rich_gini ~ protection_sts, family = beta_family(), data = tmp[[i]])
}
```

## Adjusted models kable tables

Here we do not look at Durbin Watson as when we reviewed plots of residuals they showed no temporal autocorrelation.

```{r}
#| warning: FALSE
# Disagregated moran's I
adj_spatial_temporal_tests %>%
  filter(model_name %in% c('adj_point_disag', 'adj_area_disag')) %>%
  select(model_name, habitat, moran_I) %>%
  mutate(habitat = str_to_title(habitat),
         moran_I = round(moran_I, 3)) %>%
  pivot_wider(names_from = model_name, values_from = moran_I) %>%
  rename('Type' = habitat, 
         'Point Count' = adj_point_disag, 
         'Area Search' = adj_area_disag) %>%
   kable() %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  add_header_above(c("Habitat" = 1, "Moran's Index" = 2), 
                   background = "#4A90C4", color = "white", bold = TRUE) %>%
  add_header_above(c("Disaggregated Moran's Index Statistics by Survey Method" = 3),  background = "#4A90C4", color = "white", bold = TRUE) %>% 
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, 7, 2), background = "#8EBEE2") %>%
  row_spec(seq(2, 6, 2), background = "#D6E8F5") %>%
  footnote(general = "Moran's Index values near 0 indicate no spatial autocorrelation.")%>% 
  save_kable(here('Outputs', 'figures',"adj_agg_moran.png"), zoom = 3)

# Aggregated Moran's I
adj_spatial_temporal_tests %>%
  filter(model_name %in% c('adj_point_agg', 'adj_area_agg')) %>%
  select(model_name, habitat, moran_I) %>%
  mutate(habitat = str_to_title(habitat),
         moran_I = round(moran_I, 3)) %>%
  pivot_wider(names_from = model_name, values_from = moran_I) %>%
  rename('Type' = habitat, 
         'Point Count' = 'adj_point_agg', 
         'Area Search' = 'adj_area_agg') %>%
   kable() %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  add_header_above(c("Habitat" = 1, "Moran's Index" = 2), 
                   background = "#4A90C4", color = "white", bold = TRUE) %>%
  add_header_above(c("Aggregated Moran's Index Statistics by Survey Method" = 3),  background = "#4A90C4", color = "white", bold = TRUE) %>% 
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, 7, 2), background = "#8EBEE2") %>%
  row_spec(seq(2, 6, 2), background = "#D6E8F5") %>%
  footnote(general = "Moran's Index values near 0 indicate no spatial autocorrelation.") %>% 
  save_kable(here('Outputs', 'figures',"adj_disag_moran.png"), zoom = 3)
```


## Put together results df

```{r}
adj_agg_df <- map_dfr(habitats, ~{
  m <- adj_area_agg[[.x]]
  if (is.null(m)) return(NULL)
  tidy(m) %>% mutate(habitat = .x, model = 'Area Search')
}) %>%
bind_rows(
  map_dfr(habitats, ~{
    m <- adj_point_agg[[.x]]
    if (is.null(m)) return(NULL)
    tidy(m) %>% mutate(habitat = .x, model = 'Point Count')
  })
) %>% 
  clean_names() %>% 
  filter(! grepl('sd_', term)) %>% 
  select(term, beta = estimate , se = std_error, habitat, model, p_value)

adj_disag_df <- map_dfr(habitats, ~{
  m <- adj_area_disag[[.x]]
  if (is.null(m)) return(NULL)
  tidy(m) %>% mutate(habitat = .x, model = 'Area Search')
}) %>%
  bind_rows(
    map_dfr(habitats, ~{
      m <- adj_point_disag[[.x]]
      if (is.null(m)) return(NULL)
      tidy(m) %>% mutate(habitat = .x, model = 'Point Count')
    })
  ) %>%
  clean_names() %>%
  filter(! grepl('sd_', term)) %>% 
  select(term, beta = estimate, se = std_error, habitat, model, p_value) 
```


## Create kable table of adj results

### Aggregated df
```{r}
table_data <- adj_agg_df %>%
  filter(!grepl('sd_(Intercept)|sigma', term)) %>%
  mutate(
    habitat = str_to_title(habitat),
    term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "protectionProtected" ~ "Protected",
      TRUE ~ term
    ),
    beta = round(beta, 3),
    se = round(se, 3),
    p_value = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01  ~ "< 0.01",
      p_value < 0.05  ~ "< 0.05",
      TRUE ~ as.character(round(p_value, 3))
    )
  ) %>%
  arrange(habitat, term) %>%
  pivot_wider(
    names_from = model,
    values_from = c(beta, se, p_value),
    names_glue = "{model}_{.value}"
  ) %>%
  select(habitat, term, 
         `Area Search_beta`, `Area Search_se`, `Area Search_p_value`,
         `Point Count_beta`, `Point Count_se`, `Point Count_p_value`)

# Auto generate habitat group indices
habitat_groups <- table_data %>%
  mutate(row = row_number()) %>%
  group_by(habitat) %>%
  summarise(start = min(row), end = max(row))

# Build kable
k <- table_data %>%
  select(-habitat) %>%
  rename(
    'Term' = term,
    'β' = `Area Search_beta`, 'SE' = `Area Search_se`, 'p-value' = `Area Search_p_value`,
    'β ' = `Point Count_beta`, 'SE ' = `Point Count_se`, 'p-value ' = `Point Count_p_value`
  ) %>%
  kable(align = c('l', 'r', 'r', 'r', 'r', 'r', 'r')) %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  # Model headers
  add_header_above(
    c("Model " = 1, "Area Search" = 3, "Point Count" = 3),
    background = "#4A90C4", color = "white", bold = TRUE
  ) %>%
  # Title
  add_header_above(
    c("Protection Effects on Bird Counts by Habitat and Survey Method" = 7),
    background = "#4A90C4", color = "white", bold = TRUE
  ) %>%
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, nrow(table_data), 2), background = "#8EBEE2") %>%
  row_spec(seq(2, nrow(table_data), 2), background = "#D6E8F5")

# Add habitat groups
for (i in 1:nrow(habitat_groups)) {
  k <- k %>% pack_rows(
    habitat_groups$habitat[i],
    habitat_groups$start[i],
    habitat_groups$end[i],
    background = "#4A90C4", color = "white", bold = TRUE
  )
}

k %>%
  footnote(general = "β = estimated coefficient; SE = standard error.; Both in Log Space") %>% 
  save_kable(here('Outputs', 'figures', "protection_effects_table.png"), zoom = 5)
```

### Disaggregated df

```{r}
table_data <- adj_disag_df %>%
  mutate(
    habitat = str_to_title(habitat),
    term = case_when(
      term == "factor(gap_sts)1" ~ "1",
      term == "factor(gap_sts)2" ~ "2",
      term == "factor(gap_sts)3" ~ "3",
      term == "factor(gap_sts)4" ~ "4",
      term == "factor(gap_sts)5" ~ "5",
      TRUE ~ term
    ),
    beta = round(beta, 3),
    se = round(se, 3),
    p_value = case_when(
      p_value < 0.001 ~ "< 0.001",
      p_value < 0.01  ~ "< 0.01",
      p_value < 0.05  ~ "< 0.05",
      TRUE ~ as.character(round(p_value, 3))
    )
  ) %>%
  arrange(habitat, term) %>%
  pivot_wider(
    names_from = model,
    values_from = c(beta, se, p_value),
    names_glue = "{model}_{.value}"
  ) %>%
  select(habitat, term, 
         `Area Search_beta`, `Area Search_se`, `Area Search_p_value`,
         `Point Count_beta`, `Point Count_se`, `Point Count_p_value`)

# Auto generate habitat group indices
habitat_groups <- table_data %>%
  mutate(row = row_number()) %>%
  group_by(habitat) %>%
  summarise(start = min(row), end = max(row))

# Build kable
k <- table_data %>%
  select(-habitat) %>%
  rename(
    'GAP Status'  = term,
    'β' = `Area Search_beta`, 'SE' = `Area Search_se`, 'p-value' = `Area Search_p_value`,
    'β ' = `Point Count_beta`, 'SE ' = `Point Count_se`, 'p-value ' = `Point Count_p_value`
  ) %>%
  kable(align = c('l', 'r', 'r', 'r', 'r', 'r', 'r')) %>%
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  # Model headers
  add_header_above(
    c("Model " = 1, "Area Search" = 3, "Point Count" = 3),
    background = "#4A90C4", color = "white", bold = TRUE
  ) %>%
  # Title
  add_header_above(
    c("Protection Effects on Bird Counts by Habitat and Survey Method" = 7),
    background = "#4A90C4", color = "white", bold = TRUE
  ) %>%
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, nrow(table_data), 2), background = "#8EBEE2") %>%
  row_spec(seq(2, nrow(table_data), 2), background = "#D6E8F5")

# Add habitat groups
for (i in 1:nrow(habitat_groups)) {
  k <- k %>% pack_rows(
    habitat_groups$habitat[i],
    habitat_groups$start[i],
    habitat_groups$end[i],
    background = "#4A90C4", color = "white", bold = TRUE
  )
}

k %>%
  footnote(general = "β = estimated coefficient; SE = standard error.; Both in Log Space") %>% 
  save_kable(here('Outputs', 'figures', "GAP_effects_table.png"), zoom = 5)
```

