---
title: "Focal Species Trends"
format: html
execute:
  echo: false
  warning: false
  cache: true   
editor: 
  markdown: 
    wrap: sentence
---

## Focal species models

In this workflow we will perform the following: 1.
Define our focal species from our `indicator_analysis.qmd` workflow.
The species produced in the `focal_species` dataframe were then sent to Point Blue staff scientists for further guidance on determining our focal species.
2.
Run model fit diagnostics checks.
3.
Run negative binomial GLMMM model for aggregated data where protection status is determined by protected (GAP status 1 - 2) and unprotected (GAP status 3-5).
4.
Run negative binomial GLMM model for disaggregated data with all GAP status protection levels.
5.
Summarize results with a gt summary.
6.
Visualize trends with `ggeffects` package.

## Set-Up

```{r}
#| label: set-up
# Load environment 
library(tidyverse)
library(here)
library(glmmTMB)
library(gtsummary)
library(gt)
library(sf)
library(sfarrow)
library(lmtest)
library(spdep)
library(broom.mixed)
library(ggplot2)
library(showtext)
library(marginaleffects)
library(ggeffects)

# Set seed for reproducibility
set.seed(76)

# ---- Source scripts ----

# Functions ----
source(here("R", "population_trend.R")) 
source(here("R", "autocorrelation_tests.R"))

# Theme for gt summaries ----
source(here("R", "point_blue_theme.R"))

# ---- Load in data ----
birds_joined <- st_read_parquet(here("data","data_processed","birds_joined.parquet"))

```

## Focal Species Data

```{r}
# MANUAL STEP — update from indicator analysis outputs

# Data Frame of focal species 
focal_species <- data.frame(
  habitat_type = c(
    "SHRUB",
    "HERBACEOUS",
    "HARDWOOD",
    "WATER",
    "URBAN",
    "CONIFER",
    "AGRICULTURE"
  ),
  common_name = c(
    "Wrentit",
    "Western Meadowlark",
    "Song Sparrow",
    "Western Sandpiper",
    "White-crowned Sparrow",
    "Red-breasted Nuthatch",
    "Red-winged Blackbird"
  )
)

# Apply function to extract individual species data from `birds_joined` 
focal_species_data <- population_trends(data = birds_joined, species_list = focal_species)

# Save as RDS file
saveRDS(focal_species_data, file = here("data", "data_processed", "focal_species.rds"))

```

## Diagnostic checks

In this section, model diagnostics are run to determine the most optimal model for this species subset.
The most optimal model is chosen as the model with the smallest AIC between Poisson and Negative binomial type II.
Poisson and Negative binomial are ideal for data with high levels of overdispersion.

```{r}
# Per-species family selection + dispersion diagnostics
model_diagnostics <- list()

for (sp in unique(focal_species_data$common_name)) {
  # Seperate df by species
  species_df <- focal_species_data %>% filter(common_name == sp)
  # Poisson model 
  m_pois <- glmmTMB(total_obs ~ year_scaled * protection_sts + offset(log(sample_effort)) + (1|study_area),
                    family = poisson,  data = species_df)
  m_nb2  <- glmmTMB(total_obs ~ year_scaled * protection_sts + offset(log(sample_effort)) + (1|study_area),
                    family = nbinom2, data = species_df)
  
  # Pick best model based on minimum AIC 
  best_model <- list(poisson = m_pois, nb2 = m_nb2)[[
    which.min(c(AIC(m_pois), AIC(m_nb2)))
  ]]
  #Check over dispersion with Pearson
  pearson_resid <- residuals(best_model, type = "pearson")
  n <- nobs(best_model)
  p <- length(fixef(best_model)$cond)
  # Create dataframe of model diagnositcs per species
  model_diagnostics[[sp]] <- data.frame(
    species          = sp,
    best_family      = c("poisson","nb2")[which.min(c(AIC(m_pois), AIC(m_nb2)))],
    aic_poisson      = AIC(m_pois),
    aic_nb2          = AIC(m_nb2),
    dispersion_param = sigma(best_model),
    pearson          = sum(pearson_resid^2) / (n - p)
  )
}
# Bind model diagnostics per species to one dataframe 
model_fit_df <- bind_rows(model_diagnostics)

model_fit_df
```

## Models -- Aggregated/Protection

```{r}
# Define list 
results_ag <- list()

# Loop through all species and fit models 
for (sp in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == sp)

    species_model <- glmmTMB(total_obs ~ year_scaled * protection_sts  + (1|study_area) + offset(log(sample_effort)),
                             family = nbinom2, data = species_df)
  
  results_ag[[sp]] <- species_model
}

# ---- Access model results ----
lapply(results_ag, summary)
```

### Autocorrelation test - Aggregated Model

In this section spatial and temporal autocorrelation was tested on the `results_ag` model with the Moran's Index and Durbin Watson. 

```{r}
# ---- Test Autocorrelation ----
model_data_map <- list()
for (sp in names(results_ag)) {
  set.seed(76)
  model_data_map[[sp]] <- focal_species_data %>% 
    filter(common_name == sp) %>%
    mutate(habitat_type = "all")
}
# Iterate across all models stored in results_ag
models <- lapply(results_ag, function(m) list(all = m))

# Run autocor function 
model_autocorr_results <- run_spatial_temporal_tests(models, model_data_map, habitats = "all")

model_autocorr_results

```

## Models -- Disaggregated/GAP

```{r}
# Define list 
results_disag <- list()

# Loop through all species and fit models 
for (sp in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == sp)
  
    species_model <- glmmTMB(total_obs ~ year_scaled * factor(gap_sts) + (1|study_area) + offset(log(sample_effort)),
                             family = nbinom2, data = species_df)
  
  results_disag[[sp]] <- species_model
}


# ---- Access model results ---- 
lapply(results_disag, summary)

```

### Autocorrelation test - Disaggregated Model
```{r}
# ---- Test Autocorrelation ----
model_data_map <- list()
for (sp in names(results_disag)) {
  set.seed(76)
  model_data_map[[sp]] <- focal_species_data %>% 
    filter(common_name == sp) %>%
    mutate(habitat_type = "all")
}

# Iterate across all models stored in results_disag
models <- lapply(results_disag, function(m) list(all = m))

# Run autocorrelation function 
model_autocorr_results <- run_spatial_temporal_tests(models, model_data_map, habitats = "all")

model_autocorr_results
```

## Population Trend Coefficients and Marginal Effects

In this section, population trend coefficients were extracted from each model with `broom.mixed::tidy()`.The annual unprotected trend was estimated from the `marginaleffects` package using the `avg_slopes()` function.

### Aggregated GT summary

```{r}
# Raw coefficients (protected + interaction)
raw_coef_table <- map_dfr(results_ag, ~ broom.mixed::tidy(.x, effects = "fixed", conf.int = TRUE),
                      .id = "species") %>%
  filter(term %in% c("year_scaled", "year_scaled:protection_stsunprotected")) %>%
  mutate(
    term = recode(term,
      "year_scaled" = "Protected",
      "year_scaled:protection_stsunprotected" = "Interaction"
    ),
    across(c(estimate, conf.low, conf.high), ~ round(.x, 4)),
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    )
  ) %>%
  select(species, term, estimate, conf.low, conf.high, significance) %>%
  pivot_wider(
    names_from  = term,
    values_from = c(estimate, conf.low, conf.high, significance)
  )

# Unprotected marginal trend via avg_slopes
unprotected_slopes <- map_dfr(names(results_ag), function(sp) {
  avg_slopes(results_ag[[sp]],
             variables = "year_scaled",
             by = "protection_sts",
             vcov = TRUE) %>%
    filter(protection_sts == "unprotected") %>%
    mutate(
      species = sp,
      across(c(estimate, conf.low, conf.high), ~ round(.x, 4)),
      # Redefine significance for interpretability
      significance_Unprotected = case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE ~ "ns"
      )
    ) %>%
    select(species, 
           estimate_Unprotected  = estimate, 
           conf.low_Unprotected  = conf.low, 
           conf.high_Unprotected = conf.high, 
           significance_Unprotected)
})

# Join raw coef with unprotected slopes 
raw_coef_table <- raw_coef_table %>%
  left_join(unprotected_slopes, by = "species")

n_rows <- nrow(raw_coef_table)

# Gt table 
focal_species_coef_table <- raw_coef_table %>%
  gt() %>%
  tab_header(title = "Focal Species Population Trends") %>%
  tab_spanner(label = "Year trend in protected sites",
              columns = c(estimate_Protected, conf.low_Protected,
                          conf.high_Protected, significance_Protected)) %>%
  tab_spanner(label = "Year × protection interaction",
              columns = c(estimate_Interaction, conf.low_Interaction,
                          conf.high_Interaction, significance_Interaction)) %>%
  tab_spanner(label = "Year trend in unprotected sites",
              columns = c(estimate_Unprotected, conf.low_Unprotected,
                          conf.high_Unprotected, significance_Unprotected)) %>%
  cols_label(
    species                   = "Species",
    estimate_Protected        = "β",
    conf.low_Protected        = "Lower CI",
    conf.high_Protected       = "Upper CI",
    significance_Protected    = "Sig.",
    estimate_Interaction      = "β",
    conf.low_Interaction      = "Lower CI",
    conf.high_Interaction     = "Upper CI",
    significance_Interaction  = "Sig.",
    estimate_Unprotected      = "β",
    conf.low_Unprotected      = "Lower CI",
    conf.high_Unprotected     = "Upper CI",
    significance_Unprotected  = "Sig."
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = "black", weight = "bold")),
    locations = cells_title()
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = "black", weight = "bold")),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = "black", weight = "bold")),
    locations = cells_column_spanners(everything())
  ) %>%
  tab_style(
    style = cell_fill(color = point_light_blue1),
    locations = cells_body(rows = seq(1, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_fill(color = point_light_blue2),
    locations = cells_body(rows = seq(2, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = species)
  ) %>%
  tab_source_note("Note: β = log-scale coefficient. Significance: * p<0.05, ** p<0.01, *** p<0.001, ns = not significant.")

focal_species_coef_table

# Save table 
gtsave(focal_species_coef_table, here::here("Outputs", "figures", "focal_species_coef_table.png"), zoom = 3)
```

### Disaggregated GT summary

In this section GT summaries were built only for species with significant and stable models.
Majority of species had sparse data across different GAP levels.

```{r}
# Species with stable disaggregated models
stable_disag_species <- c("Wrentit", "Western Meadowlark", "Red-winged Blackbird")

# Extract coefficients
disag_coef_table <- map_dfr(stable_disag_species, function(sp) {
  broom.mixed::tidy(results_disag[[sp]], 
                    effects = "fixed", 
                    conf.int = TRUE) %>%
    mutate(species = sp)
}) %>%
  filter(grepl("year", term)) %>%
  mutate(
    term = case_when(
      term == "year_scaled"                        ~ "Baseline (GAP 2)",
      term == "year_scaled:factor(gap_sts)3"       ~ "GAP 3",
      term == "year_scaled:factor(gap_sts)4"       ~ "GAP 4",
      term == "year_scaled:factor(gap_sts)5"       ~ "GAP 5",
      TRUE                                          ~ term
    ),
    across(c(estimate, conf.low, conf.high), ~ round(.x, 4)),
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    ),
    # Flag unstable estimates
    unstable = abs(std.error) > 100
  ) %>%
  filter(!unstable, !grepl("GAP 4", term)) %>%
  select(species, term, estimate, conf.low, conf.high, significance) %>%
  pivot_wider(
    names_from  = term,
    values_from = c(estimate, conf.low, conf.high, significance)
  )

n_rows <- nrow(disag_coef_table)

# GT summary
focal_species_coef_table_disag <- disag_coef_table %>%
  gt() %>%
  tab_header(title = "Focal Species Population Trends by GAP Status") %>%
  tab_spanner(label = "Baseline (GAP 2)",
              columns = c(`estimate_Baseline (GAP 2)`, `conf.low_Baseline (GAP 2)`,
                          `conf.high_Baseline (GAP 2)`, `significance_Baseline (GAP 2)`)) %>%
  tab_spanner(label = "GAP 3",
              columns = c(`estimate_GAP 3`, `conf.low_GAP 3`,
                          `conf.high_GAP 3`, `significance_GAP 3`)) %>%
  tab_spanner(label = "GAP 5",
              columns = c(`estimate_GAP 5`, `conf.low_GAP 5`,
                          `conf.high_GAP 5`, `significance_GAP 5`)) %>%
  cols_label(
    species                        = "Species",
    `estimate_Baseline (GAP 2)`    = "β",
    `conf.low_Baseline (GAP 2)`    = "Lower CI",
    `conf.high_Baseline (GAP 2)`   = "Upper CI",
    `significance_Baseline (GAP 2)` = "Sig.",
    `estimate_GAP 3`               = "β",
    `conf.low_GAP 3`               = "Lower CI",
    `conf.high_GAP 3`              = "Upper CI",
    `significance_GAP 3`           = "Sig.",
    `estimate_GAP 5`               = "β",
    `conf.low_GAP 5`               = "Lower CI",
    `conf.high_GAP 5`              = "Upper CI",
    `significance_GAP 5`           = "Sig."
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = "black", weight = "bold")),
    locations = cells_title()
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = "black", weight = "bold")),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = "black", weight = "bold")),
    locations = cells_column_spanners(everything())
  ) %>%
  tab_style(
    style = cell_fill(color = point_light_blue1),
    locations = cells_body(rows = seq(1, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_fill(color = point_light_blue2),
    locations = cells_body(rows = seq(2, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = species)
  ) %>%
  tab_source_note("Note: β = log-scale coefficient. Significance: * p<0.05, ** p<0.01, *** p<0.001, ns = not significant.")

focal_species_coef_table_disag

gtsave(focal_species_coef_table_disag, here::here("Outputs", "figures", "focal_species_coef_table_disag.png"), zoom = 3)

```

## Visuals
In this section, predicted population trends were generated using `ggpredict()` from the `ggeffects` package.
Predictions reflect model estimates of relative abundance over time for protected and unprotected sites.
The unprotected trend is equivalent to the marginal effect's unprotected trend.

```{r}
font_add_google("Libre Franklin", family = "libre franklin")
showtext.auto()

weme_trends_plt <- ggpredict(results_ag[["Western Meadowlark"]], 
          terms = c("year_scaled [all]", "protection_sts"),
          # Hold variables in model constant 
          condition = c(sample_effort = mean(
            focal_species_data$sample_effort[
              focal_species_data$common_name == "Western Meadowlark"]))) %>%
  as.data.frame() %>%
  mutate(year = x + min(focal_species_data$year_collected)) %>%
  ggplot(aes(x = year, y = predicted,
             color = group, fill = group)) +
  # Add uncertainty bounds
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.12, color = NA) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("protected"   = "#558731",
                                 "unprotected" = "#595959"),
                     labels = c("Protected", "Unprotected")) +
   scale_fill_manual(values  = c("protected"   = "#558731",
                                 "unprotected" = "#595959"),
                    labels  = c("Protected", "Unprotected")) +
  guides(fill = "none") +
  labs(x = "Year", 
       y = "Relative abundance",
       color = NULL,
       title = "Western Meadowlark in Grasslands",
       subtitle = "Unprotected sites increasing relative to protected over time.") +
  theme_classic(base_size = 11) +
  theme(text = element_text(family = "libre franklin"),
        plot.title = element_text(face = "italic", size = 15),
        plot.subtitle = element_text(size = 9, color = "gray40"),
        legend.position = "bottom",
        plot.background = element_rect(fill = "#e8eff7"),
        legend.background = element_rect(fill = "#e8eff7"),
        panel.background = element_rect(fill = "#e8eff7"),
        panel.grid.minor = element_blank())

weme_trends_plt 


# Save plot as png 
showtext_opts(dpi = 300)
ggsave(here("Outputs", "figures", "weme.png"), 
       plot = weme_trends_plt, 
       width = 12,
       height = 8,
       dpi = 300)

```

```{r}
sosp_trends_plot <- ggpredict(results_ag[["Song Sparrow"]], 
          terms = c("year_scaled [all]", "protection_sts"),
          # Hold variables in model constant 
          condition = c(sample_effort = mean(
            focal_species_data$sample_effort[
              focal_species_data$common_name == "Song Sparrow"]))) %>%
  as.data.frame() %>%
  mutate(year = x + min(focal_species_data$year_collected)) %>%
  ggplot(aes(x = year, y = predicted,
             color = group, fill = group)) +
  # Add uncertainty bounds
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.12, color = NA) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("protected"   = "#558731",
                                 "unprotected" = "#595959"),
                     labels = c("Protected", "Unprotected")) +
  scale_fill_manual(values  = c("protected"   = "#558731",
                                 "unprotected" = "#595959"),
                    labels  = c("Protected", "Unprotected")) +
  guides(fill = "none") +
  labs(x = NULL, 
       y = "Relative abundance",
       color = NULL,
       title = "Song Sparrow in Hardwood",
       subtitle = "Protected and unprotected sites both decreasing.") +
  theme_classic(base_size = 11) +
  theme(text             = element_text(family = "libre franklin"),
        plot.title       = element_text(face = "italic", size = 15),
        plot.subtitle    = element_text(size = 9, color = "gray40"),
         plot.background = element_rect(fill = "#e8eff7"),
        legend.background = element_rect(fill = "#e8eff7"),
        panel.background = element_rect(fill = "#e8eff7"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())

sosp_trends_plot

# Save plot as png 
showtext_opts(dpi = 300)
ggsave(here("Outputs", "figures", "sosp.png"), 
       plot = sosp_trends_plot, 
       width = 12,
       height = 8,
       dpi = 300)
```

```{r}
rwbl_trends_plt <- ggpredict(results_ag[["Red-winged Blackbird"]], 
          terms = c("year_scaled [all]", "protection_sts"),
          # Hold variables in model constant 
          condition = c(sample_effort = mean(
            focal_species_data$sample_effort[
              focal_species_data$common_name == "Red-winged Blackbird"]))) %>%
  as.data.frame() %>%
  mutate(year = x + min(focal_species_data$year_collected)) %>%
  ggplot(aes(x = year, y = predicted,
             color = group, fill = group)) +
  # Add uncertainty bounds
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high),
              alpha = 0.12, color = NA) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("protected"   = "#558731",
                                 "unprotected" = "#595959"),
                     labels = c("Protected", "Unprotected")) +
    scale_fill_manual(values  = c("protected"   = "#558731",
                                 "unprotected" = "#595959"),
                    labels  = c("Protected", "Unprotected")) +
  guides(fill = "none") +
  labs(x = NULL, 
       y = "Relative abundance",
       color = NULL,
       title = "Red-winged Blackbird in Working Lands",
       subtitle = "Protected sites maintain relatively stable populations.") +
  theme_classic(base_size = 11) +
  theme(text             = element_text(family = "libre franklin"),
        plot.title       = element_text(face = "italic", size = 15),
        plot.subtitle    = element_text(size = 9, color = "gray40"),
         plot.background = element_rect(fill = "#e8eff7"),
        legend.background = element_rect(fill = "#e8eff7"),
        panel.background = element_rect(fill = "#e8eff7"),
        legend.position  = "bottom",
        panel.grid.minor = element_blank())

rwbl_trends_plt

# Save plot as png 
showtext_opts(dpi = 300)
ggsave(here("Outputs", "figures", "rwbl.png"), 
       plot = rwbl_trends_plt, 
       width = 12,
       height = 8,
       dpi = 300)
```

