---
title: "population_trends"
format: html
---
## Focal species models
In this workflow we will perform the following:
1. Define our focal species from our `indicator_analysis.qmd` workflow. The species produced in the `focal_species` dataframe were then sent to Point Blue staff scientists for further guidance on determining our focal species. 
2. Run model fit diagnoist checks for 1) all focal species (no survey subset) 2) all focal species (survey subset). 
3. Figure out sampling effort: 1) Bin data across 2-5 years 2) quantify survey effort
3. Run GLMMM model for aggregated data where protection status is determined by protected (GAP status 1 - 2) and unprotected (GAP status 3-5). Summarize results with {gt} summary. 
4. Run GLMM model for disaggregated data with all GAP status protection levels. Summarize results with {gt} summary. 
5. Look into autocorrelation of results. 


```{r}
#| label: set-up
# Load environment 
library(tidyverse)
library(here)
library(glmmTMB)
library(gtsummary)
library(sf)
library(sfarrow)
library(lmtest)


# Source function
source(here("R", "population_trend.R")) # UPDATE

source(here("R", "autocorrelation_tests.R"))

birds_joined <- st_read_parquet(here("data","data_processed","birds_joined.parquet"))

```

### Define focal species 
```{r}
# Data Frame of focal species 
focal_species <- data.frame(
  habitat_type = c("SHRUB", "HERBACEOUS", "HARDWOOD", "WATER", "URBAN", "CONIFER", "AGRICULTURE"),
  common_name = c("Wrentit", "Western Meadowlark", "Song Sparrow", "Western Sandpiper", 
                  "White-crowned Sparrow", "Red-breasted Nuthatch", "Red-winged Blackbird")
)

focal_species <- data.frame(
  habitat_type = c("SHRUB", "HERBACEOUS", "HARDWOOD", "CONIFER", "AGRICULTURE"),
  common_name = c("Wrentit", "Western Meadowlark", "Song Sparrow", 
                  "Red-breasted Nuthatch", "Red-winged Blackbird")
)
```

```{r}
# Function reads the vector and prepares dataframe 
population_trends <- function(data, species_list) {
  data %>%
    inner_join(species_list, by = c("common_name", "habitat_type")) %>%
    group_by(common_name, year_collected, protection_sts, gap_sts, study_area, survey_type) %>%
    summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>% 
    ungroup() %>% 
    # Center years for models 
    mutate(year_scaled = year_collected - min(year_collected)) %>% 
    dplyr::select(common_name, year_collected, year_scaled, total_obs, protection_sts,gap_sts,study_area, survey_type, geometry)
}

# Apply function to data
focal_species_data <- population_trends(data = birds_joined, species_list = focal_species)


# Explore -- view obs per focal species 
explore_df <- focal_species_data %>%
  group_by(common_name) %>%
  summarize(all_obs = sum(total_obs, na.rm = TRUE))


# Where do we have data? 
years_results <- list()

for (i in unique(focal_species$common_name)) {
  habitat <- focal_species$habitat_type[focal_species$common_name == i]
  
  years_results[[i]]<- birds_joined %>%
    st_drop_geometry() %>%
    filter(common_name == i, habitat_type == habitat) %>%
    group_by(year_collected) %>%
    summarise(sum_obs = n()) %>% 
    ungroup() 
}

years_results[["Wrentit"]] # 1996 - 2024
years_results[["Western Meadowlark"]] # 1995 - 2016 
years_results[["Song Sparrow"]] # 1995 - 2024
years_results[["Western Sandpiper"]] # 1990 - 2022 
years_results[["White-crowned Sparrow"]] # 1999 - 2024  
years_results[["Red-breasted Nuthatch"]] # 1996 - 2024
years_results[["Red-winged Blackbird"]] # 1994 - 2024 

```

```{r}
# --- Subset data based on survey_type ---- 

focal_species_area <- focal_species_data %>% 
  filter(survey_type == "Area Search")

focal_species_point <- focal_species_data %>% 
  filter(survey_type == "Point Count")

# Which species are single-survey method?
single_survey_spp <- focal_species_data %>%
  group_by(common_name) %>%
  summarise(n_survey_types = n_distinct(survey_type)) %>%
  filter(n_survey_types < 2) %>%
  pull(common_name)

```

## Diagnostic checks: Model Fit -- No survey subset 
Negative binomial 1 & 2 vs. poisson 
```{r}
# ---- Models for ALL focal species ----
# Looking at which model suits all focal species first. 

# Poisson model
model_pois <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = poisson, data = focal_species_data)
# Negative binomial 1 model (linear variance to mean)
model_nb1  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = nbinom1(), data = focal_species_data)
# Negative binomial 2 model (quadratic variance to mean)
model_nb2  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = nbinom2(), data = focal_species_data)
# Compare model AIC (model fit, lower is better)
AIC(model_pois, model_nb1, model_nb2) # nbinom2 best model for ALL species

# ---- Check over dispersion with Pearson for model_nb2----
# Extract Pearson residuals
pearson_resid <- residuals(model_nb2, type = "pearson")

# Dispersion statistic
n <- nobs(model_nb2)
p <- length(fixef(model_nb2)$cond)

dispersion <- sum(pearson_resid^2) / (n - p)
dispersion


# ---- Models per focal species with dispersion parameter (nbinom2) and Pearson stat----
# Useful to see if the model won't capture variation for some species 

# Initialize lists 
model_test_results <- list()
model_diagnostics <- list()

for (i in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  model_test <- glmmTMB(total_obs ~ year_collected + (1|study_area), 
                        family = nbinom2, data = species_df)
  
  pearson_resid <- residuals(model_test, type = "pearson")
  n <- nobs(model_test)
  p <- length(fixef(model_test)$cond)
  
  model_test_results[[i]] <- model_test
  model_diagnostics[[i]] <- data.frame(
    species = i,
    dispersion_param = sigma(model_test),
    pearson = sum(pearson_resid^2) / (n - p)
  )
}
# Turn into dataframe
diagnostics_df <- bind_rows(model_diagnostics)

# Look at disperson parameters at the species level 
summary(model_test_results[["Wrentit"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Western Meadowlark"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Song Sparrow"]])
summary(model_test_results[["Western Sandpiper"]]) # high overdispersion
summary(model_test_results[["White-crowned Sparrow"]])
summary(model_test_results[["Red-breasted Nuthatch"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Red-winged Blackbird"]])


```

## Diagnostic checks: Model Fit -- Survey subset
Separate models by survey type so each survey type gets its own effort metric. Why? Area search effort is spatial and point count effort is temporal. Problem: reduces sample sizes per species. 

```{r}
# ---- Models for ALL focal species ----
# ---- Area Search ----

survey_df <- list(
  "Area Search" = focal_species_area,
  "Point Count" = focal_species_point
)

survey_diagnostics <- list()

for (i in names(survey_df)) {
  df <- survey_df[[i]]
  
  model_pois <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                    family = poisson,
                    data = df)
  model_nb1  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                    family = nbinom1(),
                    data = df)
  model_nb2  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                    family = nbinom2(),
                    data = df)
  
  survey_diagnostics[[i]] <- data.frame(
    survey_type = i,
    aic_pois = AIC(model_pois),
    aic_nb1 = AIC(model_nb1),
    aic_nb2  = AIC(model_nb2),
  )
}

survey_diagnostics <- bind_rows(survey_diagnostics)


# ---- Models per focal species with dispersion parameter (nbinom2) and Pearson stat----
# Useful to see if the model won't capture variation for some species 

# ---- Area Search -----

# Initialize lists 
model_test_results <- list()
model_diagnostics <- list()

for (i in unique(focal_species_area$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  model_test <- glmmTMB(total_obs ~ year_collected + (1|study_area), 
                        family = nbinom2, data = species_df)
  
  pearson_resid <- residuals(model_test, type = "pearson")
  n <- nobs(model_test)
  p <- length(fixef(model_test)$cond)
  
  model_test_results[[i]] <- model_test
  model_diagnostics[[i]] <- data.frame(
    species = i,
    dispersion_param = sigma(model_test),
    pearson = sum(pearson_resid^2) / (n - p)
  )
}
# Turn into dataframe
diagnostics_df <- bind_rows(model_diagnostics)

# ---- Point Count -----

# Initialize lists 
model_test_results <- list()
model_diagnostics <- list()

for (i in unique(focal_species_point$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  model_test <- glmmTMB(total_obs ~ year_collected + (1|study_area), 
                        family = nbinom2, data = species_df)
  
  pearson_resid <- residuals(model_test, type = "pearson")
  n <- nobs(model_test)
  p <- length(fixef(model_test)$cond)
  
  model_test_results[[i]] <- model_test
  model_diagnostics[[i]] <- data.frame(
    species = i,
    dispersion_param = sigma(model_test),
    pearson = sum(pearson_resid^2) / (n - p)
  )
}
# Turn into dataframe
diagnostics_df <- bind_rows(model_diagnostics)

```

# Models 
## Testing workflow: With Offset 
```{r}
# Test workflow with one species 
# ----- Aggregated -----
wren_trends_ag <- focal_species_data %>% 
  filter(common_name == "Wrentit") 

# Add site effects for study area for non-independence
# log(offset(sa))
# habitat is implicit within this model structure 
wren_model_ag <- glmmTMB(total_obs ~ year_scaled * protection_sts +
                           survey_type + offset(log(sample_effort)) +  (1|study_area),
                         family = nbinom2,
                         data = wren_trends_ag)
summary(wren_model_ag)

# ----- Disaggregated ----
wren_trends_disag <- focal_species_data %>% 
  filter(common_name == "Wrentit") %>% 
  select(year_collected, total_obs, gap_sts, study_area)

wren_model_disag <- glmmTMB(total_obs ~ year_collected * gap_sts +
                              survey_type + offset(log(sample_effort)) + (1|study_area),
                         family = nbinom2,
                         data = wren_trends_disag)
summary(wren_model_disag)


```
## Testing workflow: Without Offset 
```{r}
#--------------------------------------------------------------------------------

#-------------------------------Focal Species: Wrentit-------------------------------
# ----- Aggregated -----
wren_trends_ag <- focal_species_data %>% 
  filter(common_name == "Wrentit") 

wren_model_ag <- glmmTMB(total_obs ~ year_scaled * protection_sts +
                           survey_type +  (1|study_area),
                         family = nbinom2,
                         data = wren_trends_ag)
summary(wren_model_ag)

# ----- Disaggregated ----
wren_trends_disag <- focal_species_data %>% 
  filter(common_name == "Wrentit") 

wren_model_disag <- glmmTMB(total_obs ~ year_scaled * gap_sts +
                              survey_type + (1|study_area),
                         family = nbinom2,
                         data = wren_trends_disag)
summary(wren_model_disag)

```
## Test Autocorrelation
```{r}
# Your inputs would look like:
models <- list(wren = list(all = wren_model_ag))  # dummy habitat level
model_data_map <- list(wren = wren_trends_ag %>%
                         mutate(habitat_type = "all"))  # dummy to satisfy filter
                            
habitats <- "all"

run_spatial_temporal_tests(models, model_data_map, habitats)

```

## Models -- Aggregated/Protection 
```{r}

# Define list 
results_ag <- list()

# Loop through all species and fit models 
for (sp in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == sp)
  
  # If single-species -- run model w/o survey type
   if (n_distinct(species_df$survey_type) < 2) {
    species_model <- glmmTMB(total_obs ~ year_scaled * protection_sts + (1|study_area),
                             family = nbinom2, data = species_df)
  } else {
    species_model <- glmmTMB(total_obs ~ year_scaled * protection_sts + survey_type + (1|study_area),
                             family = nbinom2, data = species_df)
  }
  
  results_ag[[sp]] <- species_model
}


# Access model results 
summary(results_ag[["Wrentit"]]) 
summary(results_ag[["Western Meadowlark"]]) 
summary(results_ag[["Song Sparrow"]])
#summary(results_ag[["Western Sandpiper"]])
#summary(results_ag[["White-crowned Sparrow"]])
summary(results_ag[["Red-breasted Nuthatch"]])
summary(results_ag[["Red-winged Blackbird"]])

```

## Models -- Disaggregated/GAP
```{r}
# Define list 
results_disag <- list()

# Loop through all species and fit models 
for (sp in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == sp)
  
  # If single-species -- run model w/o survey type
  if (n_distinct(species_df$survey_type) < 2) {
    species_model <- glmmTMB(total_obs ~ year_scaled * gap_sts + (1|study_area),
                             family = nbinom2, data = species_df)
  } else {
    species_model <- glmmTMB(total_obs ~ year_collected * gap_sts + survey_type + (1|study_area),
                             family = nbinom2, data = species_df)
  }
  
  results_disag[[sp]] <- species_model
}


# Access model results 
summary(results_disag[["Wrentit"]]) 
summary(results_disag[["Western Meadowlark"]]) 
summary(results_disag[["Song Sparrow"]])
#summary(results_disag[["Western Sandpiper"]])
#summary(results_disag[["White-crowned Sparrow"]])
summary(results_disag[["Red-breasted Nuthatch"]])
summary(results_disag[["Red-winged Blackbird"]])

```


## Model results 

```{r}
point_count 
```


