---
title: "population_trends_trial"
format: html
---
## Purpose
This qmd is to document the trail for population trends. These trends will be assessed on placeholder species. The goal is to find matched GAP-habitat status study areas with consistent yearly data.

### Workflow 
1. Find consistent time frames of observations for placeholder species. This will be change depending on our *focal species*.
2. Determine placeholder species (randomly sample for now)
3. Fit GLMM for each habitat separately (8 models).
4. Fit GLMM for each GAP status (4 models)
3. Export as {gt} summary table.
4. Future steps:
- apply workflow to actual focal species
- Run ACF per site on residuals
- Check Moran's I for spatial clustering
- Address autocorrelation if found, then re-fit

Things that may need adjusting/implementing:
- family of model
- specification of models 
- site effects, other effects 

```{r}
#| label: set-up
# Load environment 
library(tidyverse)
library(here)
library(ggplot2)
library(sf)
library(glmmTMB)
library(gtsummary)

# Source spatial join
source(here("R", "spatial_join.R"))

# Source function
source(here("R", "population_trend.R"))

```
### Approach 1: Consistent time frames 

#### Determine consistent time frames - Testing
In order to assess population trends, we need to find consistent time frames with observations for ALL species. 
1. Diagnostic check - data across the years 
```{r}
# Observations per year
total_obs <- birds_joined %>%
  st_drop_geometry() %>% 
  # group by year 
  group_by(year_collected) %>%
  summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>%
  arrange(year_collected) 

# View observations by year 
  total_obs %>% 
    # Exclude large influxes of data 
    filter(!year_collected %in% 1988:1991) %>% 
  ggplot(aes(x = year_collected, y = total_obs)) +
  geom_line() +
  theme_classic()

```

2. Diagnostic check - data across certain sites 
Our data has 723 unique study areas. 
```{r}
study_area_obs <- birds_joined %>%
  group_by(year_collected) %>%
  st_drop_geometry() %>% 
  # Sum observations 
  summarise(
    total_obs = sum(observation_count, na.rm = TRUE),
    # sum distinct number of sites 
    n_sites = n_distinct(study_area)
  ) %>% 
  # Obs per site 
  mutate(obs_per_site = total_obs / n_sites) %>%
  arrange(year_collected) 

```
3. Diagnostic check - Sites above threshold
```{r}

year_coverage <- birds_joined %>%
  st_drop_geometry() %>% 
  filter(!year_collected %in% 1988:1991) %>%
  group_by(year_collected) %>%
  summarise(n_sites = n_distinct(study_area)) %>%
  mutate(
    pct_of_max = n_sites / max(n_sites),
    # Flag years with sufficient data 
    sufficient = pct_of_max >= 0.7
  ) 

consistent_periods <- year_coverage %>%
  filter(sufficient) %>%
  pull(year_collected)

head(consistent_periods)
```

Large influx of data from 1988-1991 due to USGS integration, maybe randomly sample points amongst those?
In summary:

- 1994:2006 and/or 1998-2005; 2019-2025

## Negative binomial vs. Poisson
```{r}

```


### Placeholder species
```{r}
set.seed(1)
# Randomly select species 
focal_species <- sample(unique(birds_joined$common_name),1)

focal_species2 <- birds_joined %>% filter(common_name == "American Robin")

# Select placeholder years based on consistent data -- this will change 
years_list <- c(1998:2005)

# Run function -- produces a subset of birds_joined with obs for focal species 
placeholder_species_trends <- population_trends(birds_joined, species_list = focal_species2, years_list = years_list)

```
## GLMM per habitat 
Goal: what is the overall effect of habitat on our species? 

Model statistical notation

$$
\begin{align}
\text{observations} &\sim \text{habitat + (1|Site)}
\end{align}
$$ 
$$
\begin{align}
\text{observations} &\sim \text{gap_sts + (1|Site) + GAP x habitat}
\end{align}
$$ 

```{r}
# .......Model 1: Habitat.....

# subset habitat
habitat_df <- unique(placeholder_species_trends$habitat_type)

# Loop through all 8 habitats and fit model 
habitat_models <- lapply(habitat_df, function(i) {
  # subset species 
  species_subset <- filter(placeholder_species_trends, habitat_type == i)
  # run model
  glmmTMB(total_obs ~ gap_sts, 
          family = nbinom2,
          data = species_subset)
})

#Subset to specific habitat /index 
summary(habitat_models)


# Can add mixed effect for site --> (1|site)
model_habitat <- glmmTMB(total_obs ~ habitat_type, 
        family = nbinom2,
        data = placeholder_species_trends)

summary(model_habitat)


```
## GLMM per GAP status
```{r}
# .......Model 2: GAP status.....

# Can add mixed effect for site --> (1|site)
model_gap <- glmmTMB(total_obs ~ gap_sts, 
        family = nbinom2,
        data = placeholder_species_trends)

summary(model_gap)
```

