---
title: "population_trends"
format: html
---
## Focal species models
In this workflow we will perform the following:
1. Define our focal species from our `indicator_analysis.qmd` workflow. The species produced in the `focal_species` dataframe were then sent to Point Blue staff scientists for furhter guidance on determining our focal species. 
2. To reduce bias from years with higher sampling effort, we will bin our data every __ year. 
3. Run GLMMM model for aggregated data where protection status is determined by protected (GAP status 1 - 2) and unprotected (GAP status 3-5). Summarize results with {gt} summary. 
4. Run GLMM model for disaggregated data with all GAP status protection levels. Summarize results with {gt} summary. 
5. Look into autocorrelation of results. 

```{r}
#| label: set-up
# Load environment 
library(tidyverse)
library(here)
library(ggplot2)
library(sf)
library(glmmTMB)
library(gtsummary)

# Source spatial join
source(here("R", "spatial_join.R"))

# Source function
source(here("R", "population_trend.R"))

```

### 1. Define focal species 
```{r}
# Data Frame of focal species 
focal_species <- data.frame(
  habitat_type = c("SHRUB", "HERBACEOUS", "HARDWOOD", "WATER", "URBAN", "CONIFER", "AGRICULTURE"),
  common_name = c("Wrentit", "Western Meadowlark", "Song Sparrow", "Western Sandpiper", 
                  "White-crowned Sparrow", "Red-breasted Nuthatch", "Red-winged Blackbird")
)
```

```{r}
# Function reads the vector and prepares dataframe 
population_trends <- function(data, species_list) {
  data %>%
    st_drop_geometry() %>%
    inner_join(species_list, by = c("common_name", "habitat_type")) %>%
    mutate(protection_sts = if_else(gap_sts %in% c("1", "2"), "protected", "unprotected")) %>% 
    group_by(common_name, year_collected, protection_sts, gap_sts, study_area) %>%
    summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>%
    ungroup() %>% 
    dplyr::select(common_name, year_collected, total_obs, protection_sts,gap_sts,study_area)
}

# Apply function to data
focal_species_data <- population_trends(data = birds_joined, species_list = focal_species)
```

## Diagnostic checks

### Negative binomial 1 & 2 vs. poisson and overdispersion
```{r}
# ---- Models for ALL focal species ----
# Looking at which model suits all focal species first. 

# Poisson model
model_pois <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = poisson, data = focal_species_data)
# Negative binomial 1 model (linear variance to mean)
model_nb1  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = nbinom1(), data = focal_species_data)
# Negative binomial 2 model (quadratic variance to mean)
model_nb2  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = nbinom2(), data = focal_species_data)
# Compare model AIC (model fit, lower is better)
AIC(model_pois, model_nb1, model_nb2) # nbinom2 best model for ALL species

# ---- Check over dispersion with Pearson for model_nb2----
# Extract Pearson residuals
pearson_resid <- residuals(model_nb2, type = "pearson")

# Dispersion statistic
n <- nobs(model_nb2)
p <- length(fixef(model_nb2)$cond)

dispersion <- sum(pearson_resid^2) / (n - p)
dispersion


# ---- Models per focal species with dispersion parameter (nbinom2) and Pearson stat----
# Useful to see if the model won't capture variation for some species 

# Initialize lists 
model_test_results <- list()
model_diagnostics <- list()

for (i in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  model_test <- glmmTMB(total_obs ~ year_collected + (1|study_area), 
                        family = nbinom2, data = species_df)
  
  pearson_resid <- residuals(model_test, type = "pearson")
  n <- nobs(model_test)
  p <- length(fixef(model_test)$cond)
  
  model_test_results[[i]] <- model_test
  model_diagnostics[[i]] <- data.frame(
    species    = i,
    dispersion_param      = sigma(model_test),
    pearson = sum(pearson_resid^2) / (n - p)
  )
}
# Turn into dataframe
diagnostics_df <- bind_rows(model_diagnostics)

# Look at disperson parameters at the species level 
summary(model_test_results[["Wrentit"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Western Meadowlark"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Song Sparrow"]])
summary(model_test_results[["Western Sandpiper"]]) # high overdispersion
summary(model_test_results[["White-crowned Sparrow"]])
summary(model_test_results[["Red-breasted Nuthatch"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Red-winged Blackbird"]])


```

### Survey effort effects
```{r}
# Choose why survey variables to use
# area-surveyed, survey_duration, sampling_unit_id, visit(?), observation_date
# I could group by year and study area to get  number of surveys/year, this won't account if there are fewer surveys with longer duration

```


### 2. Bin years
```{r}
# Investigate the year scheme per bird 
birds_joined %>%
  st_drop_geometry() %>%
  filter(common_name == "Wrentit", habitat_type == "SHRUB") %>%
  group_by(year_collected) %>%
  summarise(n = n())


# Where do we have data? 
years_results <- list()

for (i in unique(focal_species$common_name)) {
  habitat <- focal_species$habitat_type[focal_species$common_name == i]
  
  years_results[[i]]<- birds_joined %>%
    st_drop_geometry() %>%
    filter(common_name == i, habitat_type == habitat) %>%
    group_by(year_collected) %>%
    summarise(sum_obs = n()) %>% 
    ungroup() 
}

years_results[["Wrentit"]] # 1996 - 2024
years_results[["Western Meadowlark"]] # 1995 - 2016 
years_results[["Song Sparrow"]] # 1995 - 2024
years_results[["Western Sandpiper"]] # 1990 - 2022 
years_results[["White-crowned Sparrow"]] # 1999 - 2024  
years_results[["Red-breasted Nuthatch"]] # 1996 - 2024
years_results[["Red-winged Blackbird"]] # 1994 - 2024 


```

Most data for entire dataset in these habitats/focal species: shrub, herbaceous, hardwood. 

Bin years every 2 years 
```{r}
birds_joined %>%
  st_drop_geometry() %>%
  filter(common_name == "Western Meadowlark", habitat_type == "HERBACEOUS") %>%
  mutate(year_bin = floor(year_collected/2) * 2) %>% 
  group_by(year_bin) %>% 
  summarise(sum_obs = n()) %>% 
  ungroup()

```

## Testing workflow 
```{r}
# Test workflow with one species 
# ----- Aggregated -----
wren_trends_ag <- birds_joined %>% 
  st_drop_geometry() %>% 
  mutate(protection_sts = if_else(gap_sts %in% c("1", "2"), "protected", "unprotected")) %>% 
  filter(common_name == "Wrentit", habitat_type == "SHRUB") %>% 
  group_by(common_name, year_collected, protection_sts, study_area) %>%
  summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>% 
  ungroup() 

# Add site effects for study area for non-independence
# habitat is implicit within this model structure 
wren_model_ag <- glmmTMB(total_obs ~ year_collected * protection_sts + (1|study_area),
                         family = nbinom2,
                         data = wren_trends_ag)
summary(wren_model)

# ----- Disaggregated ----
wren_trends_disag <- birds_joined %>% 
  st_drop_geometry() %>% 
  filter(common_name == "Wrentit", habitat_type == "SHRUB") %>% 
  group_by(common_name, year_collected, gap_sts, study_area) %>%
  summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>% 
  ungroup() 

wren_model_disag <- glmmTMB(total_obs ~ year_collected * protection_sts + (1|study_area),
                         family = nbinom2,
                         data = wren_trends_disag)

```

## Models -- Aggregated 
```{r}

# Store  results
results_ag <- list()

# Loop through all species and run models
for (i in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  species_model <- glmmTMB(total_obs ~ year_collected * protection_sts + (1|study_area),
                            family = nbinom2,
                            data = species_df)
  
  results_ag[[i]] <- species_model
}

# Access models 
# Wrentit/Shrub -- confirm if it works 
summary(results_ag[["Wrentit"]]) 
summary(results_ag[["Western Meadowlark"]]) 
summary(results_ag[["Song Sparrow"]])
summary(results_ag[["Western Sandpiper"]])
summary(results_ag[["White-crowned Sparrow"]])
summary(results_ag[["Red-breasted Nuthatch"]])
summary(results_ag[["Red-winged Blackbird"]])

```

