---
title: "population_trends_trial"
format: html
---
## Purpose
This qmd is to document the trail for population trends. These trends will be accessed on placeholder species. 

```{r}
#| label: set-up
# Load environment 
library(tidyverse)
library(here)
library(ggplot2)
library(sf)

# Source spatial join
source(here("R", "spatial_join.R"))

```
### Approach 1: Consistent time frames 

#### Determine consistent time frames - Testing
In order to assess population trends, we need to find consistent time frames with observations for ALL species. 

1. Diagnostic check - data across the years 
```{r}
# Observations per year
total_obs <- birds_joined %>%
  st_drop_geometry() %>% 
  # group by year 
  group_by(year_collected) %>%
  summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>%
  arrange(year_collected) 

# View observations by year 
  total_obs %>% 
    # Exclude large influxes of data 
    filter(!year_collected %in% 1988:1991) %>% 
  ggplot(aes(x = year_collected, y = total_obs)) +
  geom_line() +
  theme_classic()

# Looking into na years 
na_years <- birds_joined %>% filter(year_collected == 2004:2009) %>% st_drop_geometry() 
```

2. Diagnostic check - data across certain sites 
Our data has 723 unique study areas. 
```{r}
study_area_obs <- birds_joined %>%
  group_by(year_collected) %>%
  st_drop_geometry() %>% 
  # Sum observations 
  summarise(
    total_obs = sum(observation_count, na.rm = TRUE),
    # sum distinct number of sites 
    n_sites = n_distinct(study_area)
  ) %>% 
  # Obs per site 
  mutate(obs_per_site = total_obs / n_sites) %>%
  arrange(year_collected) 
  

study_area_obs %>% 
  ggplot(aes(year_collected, obs_per_site)) +
  geom_line() +
  theme_minimal()

```
3. Diagnostic  check - Sites above threshold
```{r}

year_coverage <- birds_joined %>%
  st_drop_geometry() %>% 
  filter(!year_collected %in% 1988:1991) %>%
  group_by(year_collected) %>%
  summarise(n_sites = n_distinct(study_area)) %>%
  mutate(
    pct_of_max = n_sites / max(n_sites),
    # Flag years with sufficient data 
    sufficient = pct_of_max >= 0.7
  ) 

consistent_periods <- year_coverage %>%
  filter(sufficient) %>%
  pull(year_collected)

head(consistent_periods)
```

Large influx of data from 1988-191 due to USGS integration, maybe randomly sample points amongst those?
In summary:

- 1994:2006 and/or 1998-2005; 2019-2025

### Placeholder species
```{r}
set.seed(1)
# Randomly select species 
focal_species <- sample(unique(birds_joined$common_name),2)

# Source function
source(here("R", "population_trend.R"))

birds_joined %>% 
    st_drop_geometry() %>% 
    filter(common_name %in% focal_species,
           year_collected %in% years_list) %>% 
    group_by(common_name, year_collected) %>% 
    mutate(total_obs = sum(observation_count, na.rm = TRUE)) %>%
  ungroup() 


  select(common_name, year_collected, gap_sts, habitat_type, total_obs)

# Select placeholder years
years_list <- c(1998:2005)

#Run function
placeholder_species_trends <- population_trends(birds_joined, species_list = focal_species, years_list = years_list)

# Model 
library(MASS)

focal_species_trends <- glm.nb(total_obs ~ common_name + year_collected + gap_sts * habitat_type, 
                            data = placeholder_species_trends)

summary(focal_species_trends)

```

1. Determine placeholder species
2. Find matched GAP-habitat status study areas with consistent yearly data 
3. Run population trends 

