---
title: "population_trends_trial"
format: html
---
## Purpose
This qmd is to document the trail for population trends. These trends will be assessed on placeholder species. The goal is to find matched GAP-habitat status study areas with consistent yearly data.

### Workflow 
1. Find consistent time frames of observations for placeholder species. This will be change depending on our *focal species*.
2. Determine placeholder species (randomly sample for now)
3. Fit GLMM for each habitat separately (8 models).
4. Fit GLMM for each GAP status (4 models)
3. Export as {gt} summary table.
4. Future steps:
- apply workflow to actual focal species
- Run ACF per site on residuals
- Check Moran's I for spatial clustering
- Address autocorrelation if found, then re-fit

Things that may need adjusting/implementing:
- family of model
- specification of models 
- site effects, other effects 

```{r}
#| label: set-up
# Load environment 
library(tidyverse)
library(here)
library(ggplot2)
library(sf)
library(glmmTMB)
library(gtsummary)

# Source spatial join
source(here("R", "spatial_join.R"))

# Source function
source(here("R", "population_trend.R"))

```
### Consistent time frames 

#### Determine consistent time frames - Testing
In order to assess population trends, we need to find consistent time frames with observations for ALL species. 
1. Diagnostic check - data across the years 
```{r}
# Observations per year
total_obs <- birds_joined %>%
  st_drop_geometry() %>% 
  # group by year 
  group_by(year_collected) %>%
  summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>%
  arrange(year_collected) 

# View observations by year 
  total_obs %>% 
    # Exclude large influxes of data 
    filter(!year_collected %in% 1988:1991) %>% 
  ggplot(aes(x = year_collected, y = total_obs)) +
  geom_line() +
  theme_classic()

```
3. Diagnostic check - Sites above threshold
```{r}

year_coverage <- birds_joined %>%
  st_drop_geometry() %>% 
  filter(!year_collected %in% 1988:1991) %>%
  group_by(year_collected) %>%
  summarise(n_sites = n_distinct(study_area)) %>%
  mutate(
    pct_of_max = n_sites / max(n_sites),
    # Flag years with sufficient data 
    sufficient = pct_of_max >= 0.7
  ) 

consistent_periods <- year_coverage %>%
  filter(sufficient) %>%
  pull(year_collected)

head(consistent_periods)
```

Large influx of data from 1988-1991 due to USGS integration, maybe randomly sample points amongst those?
In summary:

- 1994:2006 and/or 1998-2005; 2019-2025


### Placeholder species
```{r}
set.seed(1)
# Randomly select species 
focal_species <- sample(unique(birds_joined$common_name),1)

# Select placeholder years based on consistent data -- this will change 
years_list <- c(1998:2005)

# Run function -- produces a subset of birds_joined with obs for focal species 
placeholder_species_trends <- population_trends(birds_joined, species_list = focal_species, years_list = years_list)

```

## Focal species models
```{r}
# Make vector of focal species 
focal_species <- c(
  "SHRUB" = "Wrentit",
  "HERBACEOUS" = "Western Meadowlark",
  "HARDWOOD" = "Song Sparrow",
  "WATER" = "Western Sandpiper",
  "URBAN" = "White-crowned Sparrow",
  "CONIFER" = "Red-breasted Nuthatch",
  "AGRICULTURE" = "Red-winged Blackbird"
)

# As a dataframe 
focal_species <- data.frame(
  habitat_type = c("SHRUB", "HERBACEOUS", "HARDWOOD", "WATER", "URBAN", "CONIFER", "AGRICULTURE"),
  common_name = c("Wrentit", "Western Meadowlark", "Song Sparrow", "Western Sandpiper", 
                  "White-crowned Sparrow", "Red-breasted Nuthatch", "Red-winged Blackbird")
)

# Test workflow with one species -- no filtering of years

wren_trends <- birds_joined %>% 
  st_drop_geometry() %>% 
  mutate(protection_sts = if_else(gap_sts %in% c("1", "2"), "protected", "unprotected")) %>% 
  filter(common_name == "Wrentit", habitat_type == "SHRUB") %>% 
  group_by(common_name, year_collected, protection_sts, study_area) %>%
  summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>% 
  ungroup() 

# Add site effects for study area for non-independence
# habitat is inplicit within this model structure 
wren_model <- glmmTMB(total_obs ~ year_collected * protection_sts,
                         family = nbinom2,
                         data = wren_trends)
summary(wren_model)


# Function reads the vector 
# Assess pop trends for ALL observation years, plan to filter to study areas with sufficient data 
population_trends <- function(data, species_list) {
  data %>%
    st_drop_geometry() %>%
    filter(common_name %in% species_list$common_name,
           habitat_type %in% species_list$habitat) %>%
    mutate(protection_sts = if_else(gap_sts %in% c("1", "2"), "protected", "unprotected")) %>% 
    group_by(common_name, year_collected, protection_sts, study_area) %>%
    summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>% 
    ungroup() %>% 
    dplyr::select(common_name, year_collected,total_obs,protection_sts, study_area)
}

# Apply function to data
focal_species_data <- population_trends(data = birds_joined, species_list = focal_species)

# Store  results
results <- list()

# Loop through all species and run models
for (i in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  species_model <- glmmTMB(total_obs ~ year_collected * protection_sts,
                            family = nbinom2,
                            data = species_df)
  
  results[[i]] <- species_model
}

# Access models 
# Wrentit/Shrub -- confirm if it works 
summary(results[["Wrentit"]]) # not the same summary 4/1/26 


```

