---
title: "population_trends"
format: html
---
## Focal species models
In this workflow we will perform the following:
1. Define our focal species from our `indicator_analysis.qmd` workflow. The species produced in the `focal_species` dataframe were then sent to Point Blue staff scientists for further guidance on determining our focal species. 
2. Run model fit diagnoist checks for 1) all focal species (no survey subset) 2) all focal species (survey subset). 
3. Run GLMMM model for aggregated data where protection status is determined by protected (GAP status 1 - 2) and unprotected (GAP status 3-5). Summarize results with {gt} summary. 
4. Run GLMM model for disaggregated data with all GAP status protection levels. Summarize results with {gt} summary. 
5. Look into autocorrelation of results. 
6. Visualize trends. 


```{r}
#| label: set-up
# Load environment 
library(tidyverse)
library(here)
library(glmmTMB)
library(gtsummary)
library(gt)
library(sf)
library(sfarrow)
library(lmtest)
library(spdep)
library(broom.mixed)
library(ggplot2)


# Source function
source(here("R", "population_trend.R")) 

source(here("R", "autocorrelation_tests.R"))

birds_joined <- st_read_parquet(here("data","data_processed","birds_joined.parquet"))

```

### Define focal species 
```{r}
# Data Frame of focal species 
focal_species <- data.frame(
  habitat_type = c(
    "SHRUB",
    "HERBACEOUS",
    "HARDWOOD",
    "WATER",
    "URBAN",
    "CONIFER",
    "AGRICULTURE"
  ),
  common_name = c(
    "Wrentit",
    "Western Meadowlark",
    "Song Sparrow",
    "Western Sandpiper",
    "White-crowned Sparrow",
    "Red-breasted Nuthatch",
    "Red-winged Blackbird"
  )
)
```

```{r}
# Function reads the vector and prepares dataframe 
population_trends <- function(data, species_list) {
  data %>%
    inner_join(species_list, by = c("common_name", "habitat_type")) %>%
    group_by(common_name, year_collected, protection_sts, gap_sts, study_area, survey_type, sample_effort, geometry) %>%
    summarise(total_obs = sum(observation_count, na.rm = TRUE)) %>% 
    ungroup() %>% 
    # Center years for models 
    mutate(year_scaled = year_collected - min(year_collected)) %>% 
    dplyr::select(common_name, year_collected, year_scaled, year_collected, total_obs, protection_sts,gap_sts,study_area, survey_type, sample_effort, geometry)
}

# Apply function to data
focal_species_data <- population_trends(data = birds_joined, species_list = focal_species)

```

## Diagnostic checks: Model Fit -- No survey subset 
```{r}
# --- Subset data based on survey_type ---- 

focal_species_area <- focal_species_data %>% 
  filter(survey_type == "Area Search")

focal_species_point <- focal_species_data %>% 
  filter(survey_type == "Point Count")

# Which species are single-survey method?
single_survey_spp <- focal_species_data %>%
  group_by(common_name) %>%
  summarise(n_survey_types = n_distinct(survey_type)) %>%
  filter(n_survey_types < 2) %>%
  pull(common_name)

```

Negative binomial 1 & 2 vs. poisson 
```{r}
# ---- Models for ALL focal species ----
# Looking at which model suits all focal species first. 

# Poisson model
model_pois <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = poisson, data = focal_species_data)
# Negative binomial 1 model (linear variance to mean)
model_nb1  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = nbinom1(), data = focal_species_data)
# Negative binomial 2 model (quadratic variance to mean)
model_nb2  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                  family = nbinom2(), data = focal_species_data)
# Compare model AIC (model fit, lower is better)
AIC(model_pois, model_nb1, model_nb2) # nbinom2 best model for ALL species

# ---- Check over dispersion with Pearson for model_nb2----
# Extract Pearson residuals
pearson_resid <- residuals(model_nb2, type = "pearson")

# Dispersion statistic
n <- nobs(model_nb2)
p <- length(fixef(model_nb2)$cond)

dispersion <- sum(pearson_resid^2) / (n - p)
dispersion


# ---- Models per focal species with dispersion parameter (nbinom2) and Pearson stat----
# Useful to see if the model won't capture variation for some species 

# Initialize lists 
model_test_results <- list()
model_diagnostics <- list()

for (i in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  model_test <- glmmTMB(total_obs ~ year_collected + (1|study_area), 
                        family = nbinom2, data = species_df)
  
  pearson_resid <- residuals(model_test, type = "pearson")
  n <- nobs(model_test)
  p <- length(fixef(model_test)$cond)
  
  model_test_results[[i]] <- model_test
  model_diagnostics[[i]] <- data.frame(
    species = i,
    dispersion_param = sigma(model_test),
    pearson = sum(pearson_resid^2) / (n - p)
  )
}
# Turn into dataframe
diagnostics_df <- bind_rows(model_diagnostics)

# Look at disperson parameters at the species level 
summary(model_test_results[["Wrentit"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Western Meadowlark"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Song Sparrow"]])
summary(model_test_results[["Western Sandpiper"]]) # high overdispersion
summary(model_test_results[["White-crowned Sparrow"]])
summary(model_test_results[["Red-breasted Nuthatch"]]) # less overdispersion, Poisson may work
summary(model_test_results[["Red-winged Blackbird"]])


```

## Diagnostic checks: Model Fit -- Survey subset
Separate models by survey type so each survey type gets its own effort metric. Why? Area search effort is spatial and point count effort is temporal. Problem: reduces sample sizes per species. 

```{r}
# ---- Models for ALL focal species ----
# ---- Area Search ----

survey_df <- list(
  "Area Search" = focal_species_area,
  "Point Count" = focal_species_point
)

survey_diagnostics <- list()

for (i in names(survey_df)) {
  df <- survey_df[[i]]
  
  model_pois <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                    family = poisson,
                    data = df)
  model_nb1  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                    family = nbinom1(),
                    data = df)
  model_nb2  <- glmmTMB(total_obs ~ year_collected + (1|study_area),
                    family = nbinom2(),
                    data = df)
  
  survey_diagnostics[[i]] <- data.frame(
    survey_type = i,
    aic_pois = AIC(model_pois),
    aic_nb1 = AIC(model_nb1),
    aic_nb2  = AIC(model_nb2),
  )
}

survey_diagnostics <- bind_rows(survey_diagnostics)


# ---- Models per focal species with dispersion parameter (nbinom2) and Pearson stat----
# Useful to see if the model won't capture variation for some species 

# ---- Area Search -----

# Initialize lists 
model_test_results <- list()
model_diagnostics <- list()

for (i in unique(focal_species_area$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  model_test <- glmmTMB(total_obs ~ year_collected + (1|study_area), 
                        family = nbinom2, data = species_df)
  
  pearson_resid <- residuals(model_test, type = "pearson")
  n <- nobs(model_test)
  p <- length(fixef(model_test)$cond)
  
  model_test_results[[i]] <- model_test
  model_diagnostics[[i]] <- data.frame(
    species = i,
    dispersion_param = sigma(model_test),
    pearson = sum(pearson_resid^2) / (n - p)
  )
}
# Turn into dataframe
diagnostics_df <- bind_rows(model_diagnostics)

# ---- Point Count -----

# Initialize lists 
model_test_results <- list()
model_diagnostics <- list()

for (i in unique(focal_species_point$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == i)
  
  model_test <- glmmTMB(total_obs ~ year_collected + (1|study_area), 
                        family = nbinom2, data = species_df)
  
  pearson_resid <- residuals(model_test, type = "pearson")
  n <- nobs(model_test)
  p <- length(fixef(model_test)$cond)
  
  model_test_results[[i]] <- model_test
  model_diagnostics[[i]] <- data.frame(
    species = i,
    dispersion_param = sigma(model_test),
    pearson = sum(pearson_resid^2) / (n - p)
  )
}
# Turn into dataframe
diagnostics_df <- bind_rows(model_diagnostics)

```

# Models 
## Testing workflow
```{r}
# Test workflow with one species 
# ----- Aggregated -----
wren_trends_ag <- focal_species_data %>% 
  filter(common_name == "Wrentit") 

wren_model_ag <- glmmTMB(total_obs ~ year_scaled * protection_sts +
                           offset(log(sample_effort)) +  (1|study_area),
                         family = nbinom2,
                         data = wren_trends_ag)
summary(wren_model_ag)

# ----- Disaggregated ----
wren_trends_disag <- focal_species_data %>% 
  filter(common_name == "Wrentit") %>% 
  select(year_collected, total_obs, gap_sts, study_area)

wren_model_disag <- glmmTMB(total_obs ~ year_collected * gap_sts +
                              offset(log(sample_effort)) + (1|study_area),
                         family = nbinom2,
                         data = wren_trends_disag)
summary(wren_model_disag)

```

## Models -- Aggregated/Protection 
```{r}
# Define list 
results_ag <- list()

# Loop through all species and fit models 
for (sp in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == sp)

    species_model <- glmmTMB(total_obs ~ year_scaled * protection_sts  + (1|study_area) + offset(log(sample_effort)),
                             family = nbinom2, data = species_df)
  
  results_ag[[sp]] <- species_model
}


# ---- Access model results ----
summary(results_ag[["Wrentit"]]) 
summary(results_ag[["Western Meadowlark"]]) 
summary(results_ag[["Song Sparrow"]])
summary(results_ag[["White-crowned Sparrow"]])
summary(results_ag[["Red-breasted Nuthatch"]])
summary(results_ag[["Red-winged Blackbird"]])
summary(results_ag[["Western Sandpiper"]])

# ---- Test Autocorrelation ----
set.seed(411)

model_data_map <- list()
for (sp in names(results_ag)) {
  model_data_map[[sp]] <- focal_species_data %>% 
    filter(common_name == sp) %>%
    mutate(habitat_type = "all")
}

models <- lapply(results_ag, function(m) list(all = m))

run_spatial_temporal_tests(models, model_data_map, habitats = "all")


```
- Song sparrow and Western Meadowlark show the most spatial autocorrelation 
## Models -- Disaggregated/GAP

```{r}
# What are the number ofobservations per species per GAp level?
gap_data <- focal_species_data %>% 
  st_drop_geometry() %>%
  count(common_name, gap_sts)

# Define list 
results_disag <- list()

# Loop through all species and fit models 
for (sp in unique(focal_species_data$common_name)) {
  species_df <- focal_species_data %>% filter(common_name == sp)
  
    species_model <- glmmTMB(total_obs ~ year_collected * factor(gap_sts) + (1|study_area) + offset(log(sample_effort)),
                             family = nbinom2, data = species_df)
  
  results_disag[[sp]] <- species_model
}


# ---- Access model results ---- 
summary(results_disag[["Wrentit"]]) 
summary(results_disag[["Western Meadowlark"]]) 
summary(results_disag[["Song Sparrow"]])
summary(results_disag[["White-crowned Sparrow"]])
summary(results_disag[["Red-breasted Nuthatch"]])
summary(results_disag[["Red-winged Blackbird"]])
summary(results_disag[["Western Sandpiper"]])

```

```{r}
raw_coef_data <- map_dfr(results_ag, ~ broom.mixed::tidy(.x, effects = "fixed", conf.int = TRUE),
                         .id = "species") %>%
  filter(grepl("year", term)) %>%
  mutate(
    term = recode(term,
      "year_scaled" = "Protected",
      "year_scaled:protection_stsunprotected" = "Unprotected"
    ),
    across(c(estimate, conf.low, conf.high), ~ round(.x, 4)),
    significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ "ns"
    )
  ) %>%
  select(species, term, estimate, conf.low, conf.high, significance) %>%
  pivot_wider(
    names_from  = term,
    values_from = c(estimate, conf.low, conf.high, significance)
  )

n_rows <- nrow(raw_coef_data)

raw_coef_table <- raw_coef_data %>%
  gt() %>%
  tab_header(title = "Focal Species Population Trends — Raw Coefficients") %>%
  tab_spanner(label = "Protected",
              columns = c(estimate_Protected, conf.low_Protected,
                          conf.high_Protected, significance_Protected)) %>%
  tab_spanner(label = "Unprotected",
              columns = c(estimate_Unprotected, conf.low_Unprotected,
                          conf.high_Unprotected, significance_Unprotected)) %>%
  cols_label(
    species                  = "Species",
    estimate_Protected       = "β",
    conf.low_Protected       = "Lower CI",
    conf.high_Protected      = "Upper CI",
    significance_Protected   = "Sig.",
    estimate_Unprotected     = "β",
    conf.low_Unprotected     = "Lower CI",
    conf.high_Unprotected    = "Upper CI",
    significance_Unprotected = "Sig."
  ) %>%
  tab_style(
    style = list(cell_fill(color = pointblue), cell_text(color = "black", weight = "bold")),
    locations = cells_title()
  ) %>%
  tab_style(
    style = list(cell_fill(color = pointblue), cell_text(color = "black", weight = "bold")),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(cell_fill(color = pointblue), cell_text(color = "black", weight = "bold")),
    locations = cells_column_spanners(everything())
  ) %>%
  tab_style(
    style = cell_fill(color = pointbluelight1),
    locations = cells_body(rows = seq(1, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_fill(color = pointbluelight2),
    locations = cells_body(rows = seq(2, n_rows, 2))
  ) %>%
  tab_source_note("Note: β = log-scale coefficient. Models fit using glmmTMB with negative binomial family. Significance: * p<0.05, ** p<0.01, *** p<0.001, ns = not significant.")

raw_coef_table

#gtsave(raw_coef_table, here::here("Outputs", "figures", "focal_species_raw_coef_table.png"), zoom = 3)
```


## Model Results
`tidy()` extracts coefficients on the log scale from the raw glmmTMB output
`exp()` converts them to rate ratios (multiplicative change per unit)
`conf.int = TRUE` adds 95% confidence intervals
```{r}
# Make function to extract coefficients
extract_coefs <- function(model_list, model_label) {
  map_dfr(model_list, ~ broom.mixed::tidy(.x, effects = "fixed", conf.int = TRUE) %>%
    mutate(across(c(estimate, conf.low, conf.high), exp),
           model = model_label),
    .id = "species")
}

# --- Extract Coefs and Metrics ---
results_tidy_ag  <- extract_coefs(results_ag,  "aggregated")

glmm_format <- function(df) {
  df %>% filter(grepl("year", term)) %>%
    mutate(
      pct_change = round((estimate - 1) * 100, 2),
      pct_low    = round((conf.low - 1) * 100, 2),
      pct_high   = round((conf.high - 1) * 100, 2),
      significance        = case_when(
        p.value < 0.001 ~ "***",
        p.value < 0.01  ~ "**",
        p.value < 0.05  ~ "*",
        TRUE            ~ "ns"
      )
    ) %>%
    select(species, model, term, pct_change, pct_low, pct_high,  significance, p.value)
}

ag_report  <- glmm_format(results_tidy_ag)

```


## Model GT summaries
```{r}
pointblue   <- "#4495D1"
pointbluelight1 <- "#A8CCE8" 
pointbluelight2 <- "#D4E6F4"  


memo_table_data <- ag_report %>%
  filter(grepl("year", term)) %>%
  mutate(
    term = recode(term,
      "year_scaled" = "Protected",
      "year_scaled:protection_stsunprotected" = "Unprotected"
    )
  ) %>%
  select(species, term, pct_change, pct_low, pct_high, significance) %>%
  pivot_wider(
    names_from  = term,
    values_from = c(pct_change, pct_low, pct_high, significance)
  ) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

n_rows <- nrow(memo_table_data)

memo_table <- memo_table_data %>%
  gt() %>%
  tab_header(title = "Focal Species Population Trends by Protection Status") %>%
  tab_spanner(label = "Protected (%/yr)",   columns = c(pct_change_Protected, pct_low_Protected, pct_high_Protected, significance_Protected)) %>%
  tab_spanner(label = "Unprotected (%/yr)", columns = c(pct_change_Unprotected, pct_low_Unprotected, pct_high_Unprotected, significance_Unprotected)) %>%
  cols_label(
    species                    = "Species",
    pct_change_Protected       = "% Change",
    pct_low_Protected          = "Lower CI",
    pct_high_Protected         = "Upper CI",
    significance_Protected     = "Sig.",
    pct_change_Unprotected     = "% Change",
    pct_low_Unprotected        = "Lower CI",
    pct_high_Unprotected       = "Upper CI",
    significance_Unprotected   = "Sig."
  ) %>%
  tab_style(
    style = list(cell_fill(color = pointblue), cell_text(color = "black", weight = "bold")),
    locations = cells_title()
  ) %>%
  tab_style(
    style = list(cell_fill(color = pointblue), cell_text(color = "black", weight = "bold")),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(cell_fill(color = pointblue), cell_text(color = "black", weight = "bold")),
    locations = cells_column_spanners(everything())
  ) %>%
  tab_style(
    style = cell_fill(color = pointbluelight1),
    locations = cells_body(rows = seq(1, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_fill(color = pointbluelight2),
    locations = cells_body(rows = seq(2, n_rows, 2))
  ) %>%
  tab_source_note("Note: Estimates are annual % change((exp(β) - 1) * 100). Models fit using glmmTMB with negative binomial family. Significance: * p<0.05, ** p<0.01, *** p<0.001, ns = not significant.")

memo_table

```

## Visuals

Overall percent change:
- visualizing song sparrow (report Moran's I), red wing blackbird, wrentit (counterintutive result), and western meadowloark (wid CIs)

```{r}
ag_report %>%
  filter(species %in% c("Song Sparrow", "Red-winged Blackbird", "Wrentit", "Western Meadowlark"),grepl("year", term)) %>%
  mutate("year_scaled" = "Protected", "year_scaled:protection_stsunprotected" = "Unprotected") %>% 
  ggplot(aes(x = pct_change, y = species, color = term)) +
  geom_pointrange(aes(xmin = pct_low, xmax = pct_high),
                  position = position_dodge(width = 0.5)) +
  labs(x = "Annual % change in population", y = NULL, color = NULL) +
  theme_minimal()

```

Raw counts

- average observation count per year 
```{r}
focal_species_data %>% filter(common_name == "Wrentit") %>%
  group_by(year_collected, total_obs) %>% 
  summarize(avg_obs = mean(total_obs, na.rm = TRUE)) %>% 
  ungroup() %>% 
  ggplot(aes(x = year_collected, y = avg_obs)) +
  geom_line()
```

```{r}
focal_species_data %>% filter(common_name == "Wrentit") %>%
  ggplot(aes(x = year_collected, y = total_obs, color = factor(gap_sts))) +
  geom_point() +
  labs(x = NULL, y = "Total Observation Count",
    color = "GAP Status",
    title = "Wrentit Population Trends by GAP Status") +
  theme_minimal()


focal_species_data %>%
  filter(common_name == "Wrentit", gap_sts %in% c(2, 4, 5)) %>%
  mutate(
    count_rate = total_obs / sample_effort,
    gap_sts    = factor(gap_sts)
  ) %>%
  group_by(year_collected, gap_sts) %>%
  summarise(mean_rate = mean(count_rate), .groups = "drop") %>%
  ggplot(aes(x = year_collected, y = mean_rate, color = gap_sts)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15) +
  facet_wrap(~ gap_sts, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = "Mean count rate (per unit effort)", title = "Wrentit Population Trends by GAP Status") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
focal_species_data %>%
  filter(common_name == "Song Sparrow") %>%
  mutate(
    count_rate     = total_obs / sample_effort,
    protection_sts = recode(protection_sts,
      "protected"   = "Protected",
      "unprotected" = "Unprotected"
    )
  ) %>%
  group_by(year_collected, protection_sts) %>%
  summarise(mean_rate = mean(count_rate), .groups = "drop") %>%
  ggplot(aes(x = year_collected, y = mean_rate, color = protection_sts)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15) +
  facet_wrap(~ protection_sts, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Protected" = "#00BFC4", "Unprotected" = "#F8766D")) +
  labs(
    x     = NULL,
    y     = "Mean count rate (per unit effort)",
    title = "Song Sparrow Population Trends") +
  theme_minimal() +
  theme(legend.position = "none")
```


