---
title: "technical_memo_figures"
format: html
---

## Load data and packages 
```{r}
# Libraries
library(sfarrow)
library(tidyverse)
library(here)
library(sf)
library(terra)
library(tigris)
library(tidyterra)
library(tmap)
library(janitor)
library(paletteer)
library(patchwork)
library(showtext)
library(kableExtra)

# Read in bird data 
birds_joined <- st_read_parquet(here('data', 'data_processed', 
                                     'birds_joined.parquet'))
# Load in habitat dataset
habitat_poly <- st_read_parquet(here('data','data_processed', 
                                     'habitat_polygon.parquet'))
# Load in GAP data
gap_clean <- st_read_parquet(here('data','data_processed', 
                                  'gap_polygon.parquet'))

# Source key functions
source(here("R", "point_blue_theme.R"))
source(here("R", "rich_gini.R"))

# Turn on style for consistently formatted plots
showtext_auto(enable = TRUE)

```

# Preliminary result plots- explore data
## Total observations over time

```{r}
# Create bar plot to show observations over time
ggplot(birds_joined, aes(x = year_collected, y = sum(observation_count, na.rm = TRUE)))+
  geom_col()+
  theme_linedraw()+
  labs(y = 'Observations', 
       x = 'Year', 
       title = 'Observations by Year')+
  point_blue_theme()
```

## Observations by year with density and area search/ point count 

```{r}
# Create a plot to display observations over time by survey type
time_obs_plot <- birds_joined %>%
  # Group by year and survey method
  group_by(year_collected, survey_type) %>%
  # Summarize obs by year and survey type
  summarise(observations = sum(observation_count, na.rm = TRUE), .groups = "drop") %>%
  # Plot 
  ggplot(aes(
    x = year_collected,
    y = observations,
    color = survey_type,
    fill = survey_type,
    group = survey_type
  )) +
  # Filled line plot
  geom_line() +
  geom_area(alpha = 0.5, position = "identity") +
  # Fill colors by survey type
  scale_fill_manual(values = c(point_blue, point_green)) +
  # Line color associated
  scale_color_manual(values = c(point_blue, point_green)) +
  # Scale y and x axes without random whitespace
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(
    expand = c(0, 0),
    # Label y axis numbers with commas to replace scientific notation
    label = scales::comma,
    limits = c(0, NA)
  ) +
  # Add labels
  labs(
    title = "Avian Knowledge Network Bird Observations Over Time",
    subtitle = "Point count data is densely clustered 1995-2005",
    y = "Count",
    x = NULL,
    fill = "Survey Type"
  ) +
  # Add point blue theme
  point_blue_theme()

# Save plot to figures folders
ggsave(
  filename = here::here("Outputs", "figures", 'time_obs_plot.png'),
  plot = time_obs_plot,
  width = 10,
  height = 8,
  dpi = 150
)

```


## Observations by GAP status

```{r}
# Create bar plot to display observations by GAP status and survey method
GAP_bar_plot <- birds_joined %>% 
  group_by(gap_sts, survey_type) %>% 
  summarise(observations = sum(observation_count, na.rm = TRUE), .groups = "drop") %>% 
ggplot(aes(x = gap_sts, y = observations, color = survey_type,
    fill = survey_type,
    group = survey_type))+
  geom_col(position = "dodge")+
  scale_fill_manual(values = c(point_blue, point_green)) +
  scale_color_manual(values = c(point_blue, point_green)) +
  scale_y_continuous(
    expand = c(0, 0),
    label = scales::comma, 
    limits = c(0, 1200000)) +
  labs(y = 'Count', 
       x = 'GAP Status', 
       title = 'Avian Observations by GAP Status', 
       subtitle = "Most observations were within GAP status 2 & 3 areas", 
         fill = "Survey Type",
  color = "Survey Type",) +
  point_blue_theme() +
  theme(axis.title.x = element_text(size = 17))+
  # Add annotation to explain NA obs
  annotate("segment", x = 6, xend = 6, y = 20000, yend = 170000) +
  annotate(geom = "text",
          x = 6, 
          y = 250000, 
          label = "observations \nin open water", 
          family = "lib_frank",
          size = 5
          )

ggsave(
  filename = here::here("Outputs", "figures", 'GAP_bar_plot.png'),
  plot = GAP_bar_plot,
  width = 10,
  height = 8,
  dpi = 150
)
```

## Observations by habitat type

```{r}
# Find and plot observations by habitat type and survey = Area Search
habitat_obs_area <- birds_joined %>% 
  filter(survey_type == "Area Search") %>% 
  group_by(habitat_type, survey_type) %>% 
  summarise(observations = sum(observation_count, na.rm = TRUE), .groups = "drop") %>% 
  mutate(habitat_type = reorder(habitat_type, observations)) %>% 
  ggplot(aes(x = observations, y = habitat_type)) +
  geom_col(fill = point_blue) +
  scale_x_continuous(label = scales::comma, expand = c(0, 0), limits = c(0, 800000)) +
  point_blue_theme() +
  theme(legend.position = "none",
        axis.title.x = element_text(size = 17)) +
  labs(x = 'Count', 
       y = 'Habitat Type', 
       title = 'Observations by Habitat in Area Searches')

# Find and plot observations by habitat type and survey = Point Count
habitat_obs_point <- birds_joined %>% 
  filter(survey_type == "Point Count") %>% 
  group_by(habitat_type, survey_type) %>% 
  summarise(observations = sum(observation_count, na.rm = TRUE), .groups = "drop") %>% 
  mutate(habitat_type = reorder(habitat_type, observations)) %>% 
  ggplot(aes(x = observations, y = habitat_type)) +
  geom_col(fill = point_blue) +
  scale_x_continuous(label = scales::comma, expand = c(0, 0), limits = c(0, 800000)) +
  point_blue_theme() +
  theme(legend.position = "none",
        axis.title.x = element_text(size = 17)) +
  labs(x = 'Count', 
       y = 'Habitat Type', 
       title = 'Observations by Habitat in Point Count')

# Combine plots using patchwork library
habitat_plot <- habitat_obs_area / habitat_obs_point

# Save plot combination
ggsave(
  filename = here::here("Outputs", "figures", 'habitat_plot.png'),
  plot = habitat_plot,
  width = 10,
  height = 8,
  dpi = 150
)
```

## County observation map
```{r}

# Pull county information for the state of California
counties <- tigris::counties() %>% 
  clean_names() %>% 
  filter(statefp == '06') %>% 
  # Transform area to be on km scale
  mutate(area = (aland+awater)* 1e-6) %>% 
  # Select only columns of interest
  select(namelsad, area) %>% 
  # Transform CRS to be the same as avian observations
  st_transform(crs(birds_joined))


# Save observation data as its own object
total_observations <- birds_joined %>% 
  select(observation_count) 
  
# Join observation data with county data using geometry
county_obs <- total_observations %>%
  st_join(counties %>% select(namelsad, area)) %>%  # join county info onto points
  # Drop bird data geometry for faster processing
  st_drop_geometry() %>%
  # Group by county
  group_by(namelsad) %>%
  # Summarize total observations at the county level
  summarize(
    total_observations = sum(observation_count, na.rm = TRUE),
    area = first(area)
  ) %>%
  # For areas without any observations, make NA = 0
  mutate(
    total_observations = replace_na(total_observations, 0),
    # Calculate density from county area by observations
    density = total_observations / area
  ) %>%
  # Arrange in data frame for easier viewing
  arrange(desc(density))

# Join counties information onto county observations
county_obs_map <- counties %>%
  left_join(county_obs, by = "namelsad")

# Create map using county boundaries to plot observation density per county
observation_map <- ggplot(county_obs_map) +
  # Using county lines (black)
  geom_sf(aes(fill = density), color = "black", linewidth = 0.2) +
  # Fill gradient by observations
  scale_fill_gradient(
  low = point_light_blue2,
  # Set upper boundary color
  high = "#1a5276",
  name = "Observations/km²",
  # Lower value color
  na.value = "grey90",
  # Color bar specifications
  guide = guide_colorbar(barwidth = 15, barheight = 0.5, title.position = "top", title.hjust = 0.5)
)+
  point_blue_theme() +
  # Hide unwanted lines and labels
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(), 
        # Put legend title on top of color bar
        legend.title = element_text(size = 17),
        # Legend is at bottom on map
        legend.position = "bottom"
        ) +
  labs(
    title = "Bird Observations Are Concentrated in Northern California",
    subtitle = "Observation density by county"
  )


ggsave(
  filename = here::here("Outputs", "figures", "observation_map.png"),
  plot = observation_map,
  width = 10,
  height = 8,
  dpi = 150
)
```



```{r}
# Create data object to do biodiversity analysis on
sts_biodiv <- birds_joined %>% 
  # Apply gini simpson function to calculate biodiversity metrics
  gini_simpson() %>% 
  # Drop geometry for faster processing
  st_drop_geometry() %>% 
  # Group by year and protection binary
  group_by(year_collected, protection_sts) %>% 
  # Mutate protection status column values for more accurate plot labels
  mutate(protection_sts = case_when(
  protection_sts == 'protected' ~ "30x30 protected",
  protection_sts == 'unprotected' ~ "non 30x30 protected"
)) %>% 
  # Summarize mean annual biodiversity for protected and non protected areas
  summarize(avg_biod = mean(rich_gini / sample_effort, na.rm = TRUE), .groups = "drop")

```

## Habitat Complexity

```{r}
# Create map of habitat areas in California 
habitats <- tm_shape(habitat_poly) +
  tm_polygons(fill = 'LIFEFORM',
              fill.scale = tm_scale(
                values = c('#B5727A',
                           '#FF72DE',
                           '#74B162',
                           '#A98858',
                           '#FEDD63',
                           '#F7A54B',
                           '#595959',
                           '#8EBEE2')),
                lty = 'blank',
              fill.legend = tm_legend_hide())+
  tm_layout(bg = FALSE, frame = FALSE,
            outer.margins = 0)

tmap_save(habitats, here('Outputs', 'figures', 'habitats.png'), 
          bg = "transparent")
```


```{r}
# Plot all 47 possible habitats
habitats_full <- rast(here('data','fveg22_1.gdb'))


habitats_whrall <- tm_shape(habitats_full)+
  tm_raster(legend.show = FALSE)+
  tm_layout(bg = FALSE, frame = FALSE,
            outer.margins = 0)

tmap_save(habitats_whrall, here('Outputs', 'figures', 'habitats_whrall.png'), 
          bg = "transparent")
```


## Gap Plots- displaying protected vs. unprotected areas in California

```{r}
# Create mask for California boundary 
california_shape <- tigris::states() %>% 
  filter(STUSPS == "CA")

bbox <- st_as_sfc(st_bbox(c(xmin = -128, xmax = -111, ymin = 29, ymax = 45), 
                          crs = st_crs(california_shape)))
mask <- st_difference(bbox, st_union(california_shape))
```

## Plot 30x30 protected areas using mask
```{r}
gap1_2 <- tm_shape(mask, bbox = st_bbox(gap_clean)) +
  tm_polygons(fill = NA, col = NA) +
tm_basemap("Esri.WorldTopoMap")+
tm_shape(gap_clean %>% 
             filter(gap_sts %in% c(1,2)))+
  tm_polygons(fill = '#73B843',
              lty = 'blank',
              col = NA) +
  tm_layout(bg = FALSE, frame = FALSE,
            outer.margins = 0)

tmap_save(gap1_2, here('Outputs', 'figures', 'gap1_2.png'), 
          bg = "transparent")
```

## Plot 30x30 unprotected areas
```{r}
gap3_4 <- tm_shape(mask, bbox = st_bbox(gap_clean)) +
  tm_polygons(fill = NA, col = NA) +
tm_basemap("Esri.WorldTopoMap")+
tm_shape(gap_clean %>% 
             filter(gap_sts %in% c(3,4)))+
  tm_polygons(fill = '#F7A54B',
              lty = 'blank',
              col = NA) + 
  tm_layout(bg = FALSE, frame = FALSE)

tmap_save(gap3_4,here('Outputs', 'figures', 'gap3_4.png'),
          bg = "transparent")
```

# Plot unknown management intent areas
```{r}
gap5 <- tm_shape(mask, bbox = st_bbox(gap_clean)) +
  tm_polygons(fill = NA, col = NA) +
tm_basemap("Esri.WorldTopoMap")+
tm_shape(gap_clean %>% 
             filter(gap_sts == 5))+
  tm_polygons(fill = charcoal,
              lty = 'blank',
              col = NA) + 
  tm_layout(bg = FALSE, frame = FALSE)

tmap_save(gap5, here('Outputs', 'figures', 'gap5.png'),
          bg = "transparent")
```

# Create basemap with no filled areas for presentation
```{r}
basemap <- tm_shape(mask, bbox = st_bbox(gap_clean)) +
  tm_polygons(fill = NA, col = NA) +
  tm_basemap("Esri.WorldTopoMap")+
   tm_layout(bg = FALSE, frame = FALSE)

tmap_save(basemap,  here('Outputs', 'figures', 'basemap.png'),
          bg = "transparent")
```


## Table of aggregated habitat categories and subcategories

```{r}
# Source geodatabase from CAL FIRE
gdb_path <- here::here('data', 'ds1327', 'tiff', 'ds1327.tif')

# Create raster object for habitat type layer
habitat_type <- rast(gdb_path)

# Grab categories
cats <- cats(habitat_type)[[1]]

# Collapse to just names and lifeform
lookup_habitat <- cats %>% 
  group_by(LIFEFORM) %>% 
  summarize(whr = paste(unique(WHRNAME), collapse = ', '))

# Create summary table
lookup_habitat %>% 
  mutate(LIFEFORM = str_to_title(LIFEFORM)) %>% 
  rename('Habitat Type' = LIFEFORM, 'Subtypes' = whr) %>% 
  kable() %>%
  column_spec(1, width = "3cm") %>% 
  kable_styling(bootstrap_options = c("hover"), full_width = F) %>%
  row_spec(0, background = "#4A90C4", color = "white", bold = TRUE) %>%
  row_spec(seq(1, 7, 2), background = "#8EBEE2") %>%
  row_spec(seq(2, 8, 2), background = "#D6E8F5") %>%
  footnote(general = "Barren Habitat dropped due to lack of data") %>% 
  save_kable(here('Outputs', 'figures',"Habitat_Types_table.png"), zoom = 3)
```


# Annual biodiversity averages model (simplified overview of protection effect and trends)
```{r}
# Simple GLM using annual biodiv averages by protection
mod_prot_main <- glm(
  avg_biod ~ year_collected + protection_sts,
  data = sts_biodiv,
  family = Gamma(link = "log")
)
summary(mod_prot_main)


# Check significance of each term
prot_coefs <- coef(summary(mod_prot_main))
year_sig <- prot_coefs["year_collected", "Pr(>|t|)"] < 0.05
prot_sig <- prot_coefs["protection_stsnon 30x30 protected", "Pr(>|t|)"] < 0.05

# Generate predictions if either term is significant
prot_preds_main <- if (year_sig | prot_sig) {
  expand.grid(
    year_collected = seq(
      min(sts_biodiv$year_collected),
      max(sts_biodiv$year_collected),
      length.out = 100
    ),
    protection_sts = unique(sts_biodiv$protection_sts)
  ) %>%
    mutate(avg_biod = predict(mod_prot_main, newdata = ., type = "response")) %>%
    filter(
      (protection_sts == "protected") |
        prot_sig  # drop unprotected line if prot not sig
    )
} else {
  tibble(
    year_collected = numeric(),
    avg_biod = numeric(),
    protection_sts = character()
  )
}

# Plot average biodiversity over time for unprotected and protected areas 
avg_biodiv_protection_plot <- ggplot(
  sts_biodiv,
  aes(
    x = year_collected,
    y = avg_biod,
    color = protection_sts,
    fill = protection_sts
  )
) +
  # Compare predictions to raw values
  geom_point(alpha = 0.5,
             size = 4,
             shape = 21) +
  {
    if (nrow(prot_preds_main) > 0)
      geom_line(
        data = prot_preds_main,
        aes(x = year_collected, y = avg_biod, color = protection_sts),
        linewidth = 1.2,
        inherit.aes = FALSE
      )
  } +
  scale_color_manual(values = c(
    "30x30 protected" = "#558731",
    "non 30x30 protected" = URBAN
  )) +
  scale_fill_manual(values = c(
    "30x30 protected" = "#558731",
    "non 30x30 protected" = URBAN
  )) +
  point_blue_theme() +
  labs(
    x = "Year",
    y = "Biodiversity Index Value",
    title = "Protected Areas Support Greater Bird Biodiversity",
    subtitle = "Average biodiversity from 1990–2025 across AKN survey sites",
    color = "Protection Status",
    fill = "Protection Status"
  )

ggsave(
  filename = here::here("Outputs", "figures", "avg_biodiv_protection_plot.png"),
  plot = avg_biodiv_protection_plot,
  width = 10,
  height = 8,
  dpi = 150
)
```

## Focal species observations
```{r}
# Load in focal species data
focal_species_data <- readRDS(here("data", "data_processed", "focal_species.rds"))

# Create observation table 
obs_table <- focal_species_data %>%
  st_drop_geometry() %>%
  group_by(common_name) %>%
  # Count number of rows/obs per species 
  summarise(
    n_protected   = sum(protection_sts == "protected"),
    n_unprotected = sum(protection_sts == "unprotected"),
    n_gap1 = sum(gap_sts == 1, na.rm = TRUE),
    n_gap2 = sum(gap_sts == 2, na.rm = TRUE),
    n_gap3 = sum(gap_sts == 3, na.rm = TRUE),
    n_gap4 = sum(gap_sts == 4, na.rm = TRUE),
    n_gap5 = sum(gap_sts == 5, na.rm = TRUE),
    total  = n(),
    .groups = "drop"
  )
# Gt summary formatting 
n_rows <- nrow(obs_table)

obs_gt_tbl <- obs_table %>%
  gt() %>%
  tab_spanner(label = "Aggregated",
              columns = c(n_protected, n_unprotected)) %>%
  tab_spanner(label = "Disaggregated (GAP Status)",
              columns = c(n_gap1, n_gap2, n_gap3, n_gap4, n_gap5)) %>%
  cols_label(
    common_name   = "Species",
    n_protected   = "Protected",
    n_unprotected = "Unprotected",
    n_gap1        = "GAP 1",
    n_gap2        = "GAP 2",
    n_gap3        = "GAP 3",
    n_gap4        = "GAP 4",
    n_gap5        = "GAP 5",
    total         = "Total"
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = point_light_blue2, weight = "bold")),
    locations = cells_column_labels(everything())
  ) %>%
  tab_style(
    style = list(cell_fill(color = point_blue),
                 cell_text(color = point_light_blue2, weight = "bold")),
    locations = cells_column_spanners(everything())
  ) %>%
  tab_style(
    style = cell_fill(color = point_light_blue1),
    locations = cells_body(rows = seq(1, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_fill(color = point_light_blue2),
    locations = cells_body(rows = seq(2, n_rows, 2))
  ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = common_name)
  ) %>%
  tab_options(
    table.border.top.style    = "hidden",
    table.border.bottom.style = "solid",
    column_labels.border.bottom.style = "solid",
    table_body.hlines.style   = "none"
  ) %>%
  tab_source_note("Note: n = number of survey records. Values of 0 indicate no observations for that GAP level.")

# View gt table of observations
obs_gt_tbl

# Save gt table
gtsave(obs_gt_tbl, 
       here::here("Outputs", "figures", "obs_summary_table.png"), 
       zoom = 3)

```


