---
title: "Indicator Analysis "
format: html
---
<<<<<<< HEAD
<<<<<<< HEAD

```{r}
=======
## Focal Species Selection
This qmd is to select focal species per habitat for population trends. 

### Workflow:
1. Run {indicspecies} package on `birds_joined` dataset (area search, point counts).
2. Join focal species with Partners in Flight.  
3. Export as csv. 

```{r}
#| label: set-up

# Load relevant packages 
>>>>>>> 5dfe080fc8e3976517b0f45ee1fc58bb32d0e2e1
=======
## Focal Species Selection
This qmd is to select focal species per habitat for population trends. 

### Workflow:
1. Run {indicspecies} package on `birds_joined` dataset (area search, point counts).
2. Join focal species with Partners in Flight.  
3. Export as csv. 

```{r}
#| label: set-up

# Load relevant packages 
>>>>>>> Nathalie
library(tidyverse) # version 2.0.0
library(sf) #1.0-21
library(here) #1.0.1
library(janitor) #2.2.1
library(stars) #0.6-8
library(terra) #1.8-42
library(arcgislayers)
library(indicspecies)
library(readxl)
<<<<<<< HEAD
<<<<<<< HEAD
```
=======
>>>>>>> Nathalie

# Source spatial join
source(here("R", "spatial_join.R"))

# Load Partners in Flight for California Bird Conservation Regions 
pif_ca <- read_excel(here("data", "ACAD Regional 2024.06.03.xlsx")) %>% 
  filter(BCR %in% c(5, 9, 15, 32, 33))

```

## Indicspecies analysis 
```{r}
#| label: dataframe-set-up 

# Prepare dataframe for multipatt
birds_indicator <- birds_joined %>%
<<<<<<< HEAD
  st_drop_geometry() %>%
=======

# Source spatial join
source(here("R", "spatial_join.R"))

# Load Partners in Flight for California Bird Conservation Regions 
pif_ca <- read_excel(here("data", "ACAD Regional 2024.06.03.xlsx")) %>% 
  filter(BCR %in% c(5, 9, 15, 32, 33))

```

## Indicspecies analysis 
```{r}
#| label: dataframe-set-up 

# Prepare dataframe for multipatt
birds_indicator <- birds_joined %>%
  st_drop_geometry() %>% 
>>>>>>> 5dfe080fc8e3976517b0f45ee1fc58bb32d0e2e1
=======
  st_drop_geometry() %>% 
>>>>>>> Nathalie
  select(study_area,
         observation_date,
         habitat_type,
         scientific_name,
         common_name, 
         observation_count) %>%
  filter(!is.na(habitat_type), !is.na(observation_count)) %>%
  group_by(study_area, observation_date, habitat_type, scientific_name) %>%
  summarise(observation_count = sum(observation_count),
            .groups = "drop") %>%
  pivot_wider(names_from = scientific_name,
              values_from = observation_count,
              values_fill = 0)

<<<<<<< HEAD
<<<<<<< HEAD
=======
# Flag any study_area with only one sample 
#birds_indicator %>% count(study_area) %>% filter(n > 1)

# Prepare group labels (habitat type) as seperate vector 
>>>>>>> Nathalie
group_vector <- birds_indicator$habitat_type

# Create numeric matrix
species_matrix <- birds_indicator %>%
select(-study_area, -observation_date, -habitat_type) %>%
  as.matrix()

```

```{r}
#| label: run-multipatt 
# Run multipatt for indicator values 
indval <- multipatt(species_matrix,
                    group_vector,
                    duleg = TRUE, # Use specificity not fidelity
                    control = how(nperm = 999)) 

<<<<<<< HEAD
=======
# Flag any study_area with only one sample 
#birds_indicator %>% count(study_area) %>% filter(n > 1)

# Prepare group labels (habitat type) as seperate vector 
group_vector <- birds_indicator$habitat_type

# Create numeric matrix
species_matrix <- birds_indicator %>%
select(-study_area, -observation_date, -habitat_type) %>%
  as.matrix()

```

```{r}
#| label: run-multipatt 
# Run multipatt for indicator values 
indval <- multipatt(species_matrix,
                    group_vector,
                    duleg = TRUE, # Use specificity not fidelity
                    control = how(nperm = 999)) 

=======
>>>>>>> Nathalie
```

```{r}
#| label: extract-indicators
<<<<<<< HEAD
>>>>>>> 5dfe080fc8e3976517b0f45ee1fc58bb32d0e2e1
=======
>>>>>>> Nathalie
# Add common names back - prevents duplicate NAs 
common_names <- birds_joined %>%
  st_drop_geometry() %>%
  distinct(scientific_name, common_name)
<<<<<<< HEAD
<<<<<<< HEAD
```
=======
>>>>>>> Nathalie

# Create indicators dataframe 
indicators <- indval$sign %>%
  rownames_to_column("species") %>%
  filter(p.value <= 0.05) %>%
<<<<<<< HEAD
  # Add common names
=======

# Create indicators dataframe 
indicators <- indval$sign %>%
  rownames_to_column("species") %>%
  filter(p.value <= 0.05) %>%
>>>>>>> 5dfe080fc8e3976517b0f45ee1fc58bb32d0e2e1
=======
>>>>>>> Nathalie
  left_join(common_names, by = c("species" = "scientific_name")) %>%
  pivot_longer(s.AGRICULTURE:s.WATER, names_to = "group", values_to = "member") %>%
  filter(member == 1) %>% 
  group_by(group) %>% 
  slice_max(stat, n = 5)
```

```{r}
<<<<<<< HEAD
<<<<<<< HEAD
# Load pif dataset 
=======
#| label: join-pif 
>>>>>>> Nathalie

# Join indicators df with pif 
focal_species <- indicators %>%
<<<<<<< HEAD
  semi_join(pif_ca, by = c("species" = "Scientific Name"))
=======
#| label: join-pif 

# Join indicators df with pif 
focal_species <- indicators %>%
=======
>>>>>>> Nathalie
  semi_join(pif_ca, by = c("species" = "Scientific Name")) %>% 
  mutate(group = str_remove(group, "s.")) 

# Add observation count to focal species 
sp_obs_habitat <- birds_joined %>%
  st_drop_geometry() %>%
  inner_join(focal_species, by = c("common_name", "habitat_type" = "group")) %>%
  group_by(common_name) %>%
  summarize(sum_obs = sum(observation_count))

focal_species <- left_join(focal_species, sp_obs_habitat, by = "common_name")

# Clean up dataframe 

focal_species <- focal_species %>% 
  select(-c(index, stat, p.value, member)) %>% 
  rename(habitat_type = group)

# Write csv 
focal_species <- write_csv(focal_species, "focal_species_selection.csv")

test <- read_csv(here("Analysis", "focal_species_selection.csv"))
<<<<<<< HEAD
>>>>>>> 5dfe080fc8e3976517b0f45ee1fc58bb32d0e2e1
=======
>>>>>>> Nathalie

```


